A stable microtubule bundle formed through an orchestrated multistep process controls quiescence exit

Version 1

Abstract

Abstract

Cells fine-tune microtubule assembly in both space and time, to give rise to distinct edifices with specific cellular functions. In proliferating cells, microtubules are highly dynamics, yet, proliferation cessation often lead to their stabilization. One of the most stable microtubule structures identified to date is the nuclear bundle assembled in yeast quiescent cells. In this report, we characterize the original multistep process driving the assembly of this structure in an AuroraB/Ipl1-dependent mechanism. This process follows a precise temporality that relies on the sequential actions of kinesin-14, kinesins-5 and involves both microtubule-kinetochore and kinetochore-kinetochore interactions. Upon quiescence exit, the microtubule bundle disassembles via a cooperative process involving the Kinesin-8 and its full disassembly is required to authorize cells re-entry into proliferation. Overall, our study not only provides the first description, at the molecular scale, of the entire life cycle of a stable microtubule structure in vivo, but also sheds light on its function as a sort of “checkpoint” for cell cycle resumption.

Introduction

Microtubules (MT) are hollow cylindrical polymers assembled by the non-covalent interaction of α- and β-tubulin heterodimers. MTs are generally nucleated at MT organizing centers (MTOC) by a ɣ-tubulin complex (ɣ-TuC) that acts as a template and stabilizes the so-called MT “minus-end” (Liu et al., 2021; Roostalu and Surrey, 2017; Sanchez and Feldman, 2017; Thawani and Petry, 2021). At the opposite end, the “plus-end” (+end), MTs elongate by the addition of GTP tubulin. The β-tubulin bound GTP is hydrolyzed and stable but transient GDP + Pi intermediates are generated in the region immediately trailing the growing +end. A subsequent Pi release favors MT depolymerization that can be rescued by de novo GTP-tubulin addition (Cleary and Hancock, 2021; Gudimchuk and McIntosh, 2021). As such, MTs alternate periods of growth and shrinkage, a behavior called dynamic instability (Mitchison and Kirschner, 1984). In vivo, a profusion of MT associated proteins (MAPs) regulate MT length and dynamics (Bodakuntla et al., 2019; Goodson and Jonasson, 2018). Besides, MTs are often assembled from multiple tubulin variants or isotypes, and modified by a cohort of post-translational modifications that modulate MT dynamics directly or by influencing the recruitment of MAPs (Janke and Magiera, 2020; Roll-Mecak, 2020). Other MAPs organize MTs into high-ordered edifices either by cross-linking MTs or by connecting them with cellular structures such as membranes, chromosomes or components of the actin cytoskeleton (Bodakuntla et al., 2019; Meiring et al., 2020). Cells fine-tune these mechanisms both in space and in time to give rise to distinct MT architectures with specific function.

In proliferating cells, MT dynamic is crucial for their functions. It allows exploring the cell volume searching for structures to “capture”, like centromeres during mitosis, or to exert pushing or pulling forces required for various cellular processes such as cell migration or morphogenesis (Heald and Khodjakov, 2015; Kirschner and Mitchison, 1986). Yet, the MT cytoskeleton can extensively rearrange and assemble more stable MT structures, notably when cells change fate (Meiring et al., 2020; Röper, 2020). For example, in terminally-differentiated cells, such as epithelial cells, cardiomyocytes or neurons, stable MT networks make up the majority, and ensure critical cellular functions such as cell shape maintenance or long-distance intra-cellular transport (Baas et al., 2016; Muroyama and Lechler, 2017). Defects in MT stabilization are at the origin of many human pathologies, including neuro-degenerative diseases and ciliopathies (Anvarian et al., 2019; Wheway et al., 2018). Effectors responsible for MT stabilization have been actively searched for since the 1980’s. In mammals, the involvement of specific MAPs such as Tau, MAP2, MAP6 and PRC1, or tubulin post-translational modifications have been extensively studied. While their contributions to MT stabilization are central and undisputable, their sole actions do not fully explain the observed levels of MT stability in many types of non-dividing cells (Hahn et al., 2019).

For years, yeast species have been instrumental in deciphering mechanisms that regulate MT dynamics in eukaryotes. While proliferating yeast cells display a dynamic MT network (Winey and Bloom, 2012), proliferation cessation goes with the formation of dramatically stable MT structures in both S. cerevisiae and S. pombe (Laporte and Sagot, 2014; Laporte et al., 2013, 2015). Upon quiescence establishment, S. cerevisiae assemble a bundle composed of stable parallel MTs, hereafter called Q-nMT bundle, for Quiescent-cell nuclear Microtubule bundle, that emanates from the nuclear side of the spindle pole body (SPB), the yeast equivalent of the centrosome. The Q-nMT bundle spans the entire nucleus and relocalizes kinetochores and centromeres that remain attached to its MT +ends (Laporte and Sagot, 2014). Since MTs embedded into the Q-nMT bundle are not all of the same length, they confer to the structure a typical arrow shape. When cells exit quiescence, the Q-nMT bundle depolymerizes and, by pulling the attached centromeres back to the SPB, allows the recovery of the typical Rabl-like configuration of chromosomes found in G1 yeast cells (Jin et al., 1998). The molecular mechanisms underlying the formation of this peculiar stable MT structure and its physiological function(s) are not understood. Yet, cells impaired for Q-nMT bundle formation have a compromised survival rate in quiescence and a drastically reduced fitness upon cell cycle re-entry (Laporte and Sagot, 2014; Laporte et al., 2013, 2015).

Here, we demonstrate that the Q-nMT bundle formation is a multistep process that follows a precise temporal sequence. The first step requires the kinesin-14 Kar3, its regulator Cik1 and the EB1 homolog Bim1, and leads to the formation of a short (≈ 1 µm) and stable bundle that resembles a half mitotic spindle. Importantly, in this first step, MT polymerization and stabilization are coupled. In a second step, additional MTs polymerize from the SPB, elongate, and are zipped to pre-existing ones in a Cin8/kinesin-5 dependant manner. The full Q-nMT bundle stabilization is established in a third and last step that requires the kinesin-5 Kip1. Our observations further indicate that Q-nMT bundle assembly requires not only MT-kinetochore attachment but also inter-kinetochore interactions. Finally, we show that, upon quiescence exit, the Q-nMT bundle disassembles from its +ends, where each MT depolymerizes in coordination with its neighbors, via the action of the depolymerase Kip3, a member of the kinesin-8 family. A complete Q-nMT bundle disassembly is required to authorize cells re-entry into the proliferation cycle. Overall, this study describes for the first time the entire life cycle of a MT structure from the molecular mechanisms involved in its formation and stabilization to its disassembly and further reveals that this atypical quiescent specific structure acts as a sort of “checkpoint” for cell cycle resumption upon quiescence exit.

Results

The Q-nMT bundle formation is a three-step process

To investigate whether the Q-nMT bundle was stable by default, i.e. intrinsically stable, or assembled as a dynamic structure and then stabilized, we quantified Q-nMT bundle length (Fig. 1A) and thickness (Fig. 1B) upon quiescence establishment following carbon source exhaustion. We found that the Q-nMT bundle assembly was a 3-step process. In an initial phase (phase I), MTs elongated from the SPB to reach ≈0.8 µm. The number of these so-called phase I-MTs was about the same as in a mitotic spindle (see inset of Fig. 1B and Sup. Fig. 1A). Importantly, phase I-MTs are resistant to nocodazole (Noc), a MT poison that causes dynamic MT depolymerization, indicating that their stabilization was concomitant with their polymerization (Fig. 1A-B). In a second phase, that started ≈10h after glucose exhaustion, additional MTs emerged from the SPB (phase II-MTs) and elongated along phase-I MTs, approximatively doubling the thickness of the phase I MT bundle (Fig. 1B and Sup Fig. 1A). At this stage, the tip of the newly elongated MTs were instable (Fig. 1A-B). Full Q-nMT bundle stabilization was reached ≈48h after glucose exhaustion (phase III), as Noc treatment affected neither Q-nMT bundle length nor its thickness (Fig. 1A-B and Sup. Fig. 1B). Another way to shed light on the Q-nMT bundle formation steps was to plot Q-nMT bundle width as a function of length for each individual cells before and after Noc treatment (Fig. 1C).

The formation of the Q-nMT bundle is a three-step process.

(A) Nuclear MT length in WT cells expressing mTQZ-Tub1, before (grey) or after a 15 min Noc treatment (blue) upon entry into quiescence.

(B) MT fluorescence intensity was used as a proxy of Q-nMT bundle width in WT cells expressing mTQZ-Tub1; thin line: intensity of an individual cell; bold line: mean intensity; n>60 for each phase. Mean intensity measurement for half pre-anaphase mitotic spindles (purple) phase I (green), phase II (orange) or phase III (red) Q-nMT bundle. To help comparison, in each graph, the dash line indicate the mean intensity in half pre-anaphase mitotic spindle. Representative cells are shown (images in pseudo-colors).

(C) Morphometric Q-nMT bundle property distribution (length and width) in each phase before and after a 15 min Noc treatment in WT cells expressing mTQZ-Tub1.

(D) Single WT cells expressing mTQZ-Tub1 (red) and Nuf2-GFP (green) in phase II (23h) or phase III (50h) were deposited on an agarose pad containing Noc and imaged. Blue arrowheads: SPB; white arrows: Nuf2-GFP clusters; time is in min after pad deposition.

(E) Tub4-mTQZ fluorescence intensity measured at the SPB upon entry into quiescence. Representative cells are shown (images in pseudo-colors). Bar is 1µm

(F) WT cells expressing mTQZ-Tub1 (red) under the TUB1 promoter and mRuby-Tub1 (green) under the ADH2 promoter. Representative cells at the indicated time after glucose exhaustion and the associated line-scans are show. The graph shows the percentage of cells harboring both mTQZ and mRuby fluorescence along the Q-nMT bundle. Each circle is the percentage obtained for an independent cell culture, n>200).

(G) Schematic of the Q-nMT bundle formation. During phase I, stable MTs (green) elongate from the SPB (in grey). During phase II, new MTs (orange) elongated from the SPB and are stabilized along the phase I MTs, but stay dynamic when longer than phase I MTs. In the meantime, Tub4 (blue) increases at the SPB (grey). After phase III, all MTs are stable (red).

(H) Upon glucose exhaustion, WT cells expressing mTQZ-Tub1 (green) and Nuf2-GFP (red) were pulsed treated with Noc (blue) or DMSO (grey) for 24 h. Noc or DMSO were then chased using carbon exhausted medium and cells were imaged. Representative images of cells 2 d after the chase are shown. Right panel: same experiment done in WT prototroph cells and representative images of cells 4 d after the chase are shown. Tubulin (green) was detected by immunofluorescence, actin (red) by phalloidin and DNA (blue) with DAPI. The mean Q-nMT bundle length (±SD) in the population is indicated.

In A, C, E an H, each circle corresponds to a single cell. In A, E, F and H mean and SD are shown, and unpaired T-test p-values are indicated; ***: p-value<10-5. In A, D, F and H bar is 2µm.

To confirm the above multistep process, we first followed Nuf2-GFP, a protein that localize at the MT +end. In phase I, Nuf2-GFP signal relocates from SPB to the distal part of elongating Q-nMT bundles (Sup. Fig. 1D). In phase II, when cells were treated with Noc, the Nuf2-GFP signal followed depolymerizing MTs (Fig. 1D), testifying for the instability of phase II MTs. By contrast, the Nuf2-GFP signal remained immobile in phase III, when cells have fully stabilized Q-nMT bundles (Fig. 1D). Second, we measured the γ-tubulin (Tub4) signal at the SPB and found that Tub4 started accumulating at the SPB at the onset of phase I to reach a plateau at the end of phase II (Fig. 1E and Sup. Fig. 1E). Interestingly, the amount of Tub4 at the SPB was proportional to the thickness of the Q-nMT bundle (Sup. Fig. 1F). Finally, in order to confirm the two waves of MT elongation, we developed a strain expressing mTQZ-Tub1 from the endogenous TUB1 promoter and mRuby-TUB1 under the ADH2 promoter, a promoter that is active only after glucose exhaustion. We observed that mRuby-TUB1 was incorporated into the Q-nMT bundle only after phase I (Fig. 1F and Sup. Fig. 1G) confirming the existence of a second wave of MT elongation in phase II. All the above information lead us to propose the model presented in Fig. 1G (see figure legend).

Q-nMT bundle formation is an active process regulated in time

Increasing viscosity of the cytoplasm was shown to dampen MT dynamics (Molines et al., 2022). Upon quiescence establishment, yeast cells experience a fluid phase to solid-like transition (Munder et al., 2016; Joyner et al., 2016) and an acidification (Jacquel et al., 2021) that could contribute to Q-nMT bundle formation. Several evidences indicate that these changes in physico-chemical properties are not involved in Q-nMT formation. First, 4-day-old quiescent cells can simultaneously display both dynamic cytoplasmic MTs (cMT) and a stable Q-nMT bundle (Sup. Fig. 1H and (Laporte et al., 2013)). Second, when we mimicked a fluid phase to solid like transition in proliferating cells by artificially dropping the pH, no stable MTs were observed, while F-actin aggregation was induced (Sup. Fig. 1I) as expected from (Peters et al., 2013). Third, 2-day-old quiescent cells displaying a phase III Q-nMT bundle were able to grow de novo dynamic MTs after a pulsed-chase Noc treatment (Sup. Fig. 1J).

If Q-nMT bundle formation was solely depending on physico-chemical changes, this structure should be able to assemble in late quiescence. By treating cells with Noc upon glucose exhaustion, we managed to prevent Q-nMT bundle formation in early quiescence (Fig. 1H). Strikingly, when the drug was washed out 24 h later, no Q-nMT bundle assembled, even several days after Noc removal (Fig. 1H). We verified that these cells did enter quiescence as they assembled actin bodies (Fig. 1H, right panel), another quiescent cell specific structure (Sagot et al., 2006). This experiment strongly suggests that Q-nMT formation is an active process induced by a transient signal emitted upon glucose exhaustion.

Mutants that cannot assemble Q-nMT bundle have a reduced viability in quiescence and a decreased capacity to form colonies upon quiescence exit (Laporte et al., 2013). To establish a causal link between the absence of Q-nMT bundle and the above phenotypes, we took advantage of our ability to prevent conditionally Q-nMT bundle formation in WT cells using early Noc treatment. We demonstrated that in the absence of Q-nMT bundle, both prototroph and auxotroph WT cells loose viability in quiescence and that survivors had a reduced capacity to generate a progeny upon quiescence exit (Sup. Fig. 1K). This reinforce the idea that the Q-nMT bundle is required for cell survival upon chronological aging.

Tubulin steady state level and Tub3 isoform are critical for Q-nMT bundle formation

We then focused on deciphering the molecular mechanisms involved in Q-nMT bundle formation. Yeast cells express two α-tubulin isotypes, Tub1 and Tub3. In proliferating cells, Tub1 make up the majority (Aiken et al., 2019; Nsamba et al., 2021; Schatz et al., 1986). In 4-day-old cells, both isoforms were embedded into the Q-nMT bundle (Fig. 2A and Sup. Fig. 2A). As Tub3 stabilizes MTs in vitro (Bode et al., 2003), it may be key for Q-nMT bundle formation in quiescence. To test this hypothesis, we first analyzed the phenotype of tub3Δ cells. In phase II, tub3Δ cells displayed short MT bundles that were highly sensitive to Noc. In phase III, no MT bundle could be detected (Fig. 2B). Thus in this mutant, MT bundles assembled during phase I were not stable and ultimately collapsed. Yet, in tub3Δ cells, the steady state amount of α-tubulin is significantly decreased (Sup. Fig. 2B and (Nsamba et al., 2021)). Thus, the amount of α-tubulin, but not Tub3 itself, could be important for Q-nMT bundle formation. Indeed, mutants impaired for α- and β-tubulin folding, such as the prefolding complex mutants pac10Δ, gim3Δ or yke2Δ, or for tubulin heterodimer formation such as pac2Δ, or for β-tubulin folding, such as cin2Δ or cin4Δ, all of which displaying a decreased amount of tubulin, were unable to assemble a Q-nMT bundle (Sup. Fig. 2C-D).

Q-nMT bundle formation is influenced by the alpha-tubulin amount and isoform

(A) WT cell (4 d) expressing either Tub3-3GFP (green) and mTQZ-Tub1 (red, top panel) or Tub3-RFP (red) and mWasabi-Tub1 (green, bottom panel). Blue arrowheads point to SPB.

(B) Nuclear MT length in WT and tub3Δ cells expressing mTQZ-Tub1, 36 h (phase II, yellow) and 90 h (phase III, red) after glucose exhaustion, treated 15 min (blue) or not (grey) with Noc. Representative images are shown on the right.

(C) Nuclear MT length in WT and Tub1-only cells expressing mTQZ-Tub1, 36 h (phase II, yellow) and 90 h (phase III, red) after glucose exhaustion, treated 15 min or not with Noc. Representative images are shown on top.

(D) Fluorescence intensity along the Q-nMT bundle in WT (grey) and Tub1-only (blue) cells expressing mTQZ-Tub1 grown for 10 h (phase I), 36 h (phase II), and 90 h (phase III). Thin lines indicate individual cells fluorescence intensity and bold lines, the mean intensity. Dashed lines indicate the maximal mean fluorescence intensity measured in phase I.

In B and C, each circle corresponds to a single cell. MT mean length, SD, and unpaired T-test p-value are indicated. In all panels, bar is 2µm.

To identify the role of Tub3 per se, we took advantage of the “Tub1-only” mutant in which the TUB3 gene was replaced by the TUB1 gene at the TUB3 locus. This mutant only expresses Tub1 and displays an α-tubulin level similar to WT (Sup. Fig. 2B and (Nsamba et al., 2021)). We found that “Tub1-only” cells display shorter but stable Q-nMT bundles in phase III (Fig. 2C). Interestingly, “Tub1-only” Q-nMT bundles were thinner than WT Q-nMT bundles (Fig. 2D). Overall, these data indicate Tub3 is involved in MT elongation in phase II.

Kinetochores are critical for phase I but dispensable for the maintenance of Q-nMT bundles

Kinetochore components are enriched at the Q-nMT bundle +end ((Laporte et al., 2013) and Sup Fig. 3A). We thus wondered if kinetochore-MT attachment could play a role in Q-nMT bundle formation. First, we disrupted kinetochore-MT attachment upon phase I using the well-established ndc80-1 allele (Cheeseman et al., 2006; DeLuca et al., 2018; Wigge et al., 1998). As shown in Fig. 3A, no Q-nMT bundles were detected in ndc80-1 cells shifted to 37°C at the onset of quiescence entry. Second, we focused on the chromosomal passenger complex (CPC: Bir1/Survivin, Sli15/INCENP, Nbl1/Borealin and Ipl1/Aurora B), a complex known to regulate kinetochore-MT attachment dynamics (Cairo and Lacefield, 2020). We found that upon quiescence entry, Ipl1 inactivation impeded MT bundle formation (Fig. 3B). While we could not detect Ipl1 and Ndc1 in quiescent cells, we found that Sli15-GFP and Bir1-GFP localized along the Q-nMT bundle with an enrichment at the Q-nMT bundle +end (Fig. 3C). The deletion of Sli15 did not affect Q-nMT bundle (Fig. Sup. 3E). However, in 4-days old bir1Δ cells, Q-nMT bundle length were slightly shorter and not fully stabilized (Fig. 3D). Finally, we investigated the involvement of Bim1, a MT +end binding protein that plays a role in kinetochore MT-end on attachment (Dudziak et al., 2021; Thomas et al., 2016). We have previously shown that Bim1 localizes all along the Q-nMT bundle but that its deletion has no impact on Q-nMT bundle formation even in the presence of Noc ((Laporte et al., 2013) and Fig. 3D). Interestingly, the amount of Bim1 was correlated with tubulin incorporation during Q-nMT bundle formation (Sup Fig. 3D). When we combined bim1Δ with bir1Δ, cells hardly assembled MT structures that vanished after Noc treatment (Fig. 3D). This demonstrates that in the absence of Bim1, Bir1 was critical for Q-nMT bundle stabilization in phase I. A similar phenotype was observed with SLI15 deletion (Sup Fig. 3E). Interestingly, the viability of bim1Δ bir1Δ cells in quiescence was drastically reduced (Sup Fig. 3G). Together, these experiments show that kinetochore-MT attachments are critical for the initiation of Q-nMT bundle assembly.

Kinetochore-kinetochore interactions are required for Q-nMT bundle formation.

(A) Nuclear MT length distribution in WT (grey) and ndc80-1 (violet) cells expressing mTQZ-Tub1 (green) and Nuf2-GFP (red), shifted to 37 °C upon glucose exhaustion for the indicated time and imaged after a 20 min Noc treatment. Representative cells shifted for 12 h at 37°C are shown.

(B) WT (grey) and ipl1-1 cells (pink) expressing mTQZ-Tub1 were shifted upon glucose exhaustion to 37 °C for 48 h, and imaged after a 20 min Noc treatment. Representative cells are shown.

(C) WT cells (2 d) expressing mTQZ-Tub1 (red) and Bir1-GFP or Sli15-GFP (green) were imaged. Graphs show Bir1-GFP or Sli15-GFP fluorescence intensity along normalized Q-nMT bundles (plain and dash lines: mean and SD respectively).

(D) Nuclear MT length distribution in cells of the indicated genotype (4 d) expressing mTQZ-Tub1 treated or not with Noc. Representative cells are shown.

(E) WT cells expressing mTQZ-Tub1 (red) and Slk19-GFP (green) 4 h after glucose exhaustion. Blue arrowhead: SPB.

(F) Nuclear MT length distribution in cells of the indicated genotype (4 d) expressing mTQZ-Tub1 (green) and Nuf2-GFP (red) were imaged before or after Noc treatment. Representative cells are shown. White arrowheads point to Nuf2 dots.

(G) Cells of the indicated genotype (4 d) expressing mTQZ-Tub1 (green) and Nup2-RFP (red).

(H) Nuclear MT length distribution in cells of the indicated genotype (4 d) expressing mTQZ-Tub1 treated or not with Noc. Representative cells are shown.

(I) Length variation of nuclear MT bundle fragments after laser ablation (pink dash line) in cells expressing mRuby-TUB1 (red) and Dad2-GFP (green). Time is in min. Blue arrowhead: SPB, white arrowhead: cMT.

In A, B, D, F, and H, each circle corresponds to a single cell. Mean, SD, and unpaired T-test p-values are indicated (*: p-value <0.05, and ***: p-value <10-6). Bar is 2 µm except in (I) where it is 1 µm.

We then hypothesized that inter-kinetochore interactions may stabilize the parallel MTs embedded into the Q-nMT bundle. To test this idea, we analyzed MT organization in slk19Δ cells that are defective in kinetochore clustering (Richmond et al., 2013). In quiescence, Slk19 was enriched at both Q-nMT bundle extremities (Fig. 3E). slk19Δ cells displayed shorter and thinner MT structures that were not stable, i.e. sensitive to Noc treatmant (Fig. 3F). Since Bim1 has been involved in kinetochore-kinetochore interaction, we combined slk19Δ with bim1Δ. MT structures detected in slk19Δ bim1Δ cells hardly reached ≈1 µm and vanished upon Noc treatment. In addition, kinetochores localized as a rosette around the SPB (Fig. 3F) and cell viability was strongly compromised (Sup. Fig. 3G). This demonstrates that in the absence of Bim1, Slk19 is strictly essential for MT bundling and stabilization in phase I. Furthermore, since kinetochores are known to be cross-linked by the monopolin complex (Mam1, Lrs4, Hrr25 and Csm1 (Corbett et al., 2010)) we analyzed the phenotype of mam1Δ, lrs4Δ and csm1Δ cells in quiescence. Most of them displayed short MT structures often arranged in a star-like array (Fig. 3G-H, Sup. Fig. 3F). As for Bir1 and Skl19, the combination of bim1 deletion with monopolin disruption worsens MT structure length and stability (Fig. 3H) and the cell viability was strongly compromised (Sup. Fig. 3G). Cells impaired for Spo13, a protein involved in monopolin recruitment to kinetochores (Lee et al., 2004) displayed a similar phenotype (Fig. 3G-H). Altogether, these experiments show that inter-kinetochore interactions are critical for Q-nMT bundle assembly.

Lastly, we investigated if kinetochore-MT interactions were required for Q-nMT bundle stability once it is formed. In 5-days old WT cells, we used a UV-pulsed-laser to break the Q-nMT bundle into two pieces (Fig. 3I). The presence of dynamic cMTs after laser ablation testified for cell viability (Fig. 3I). In most cases, the length of both the released fragment and the fragment attached to the SPB remained constant (Fig. 3I, Sup. Fig. 3H). By contrast, in proliferating cells, dynamic anaphase spindles promptly disassembled following a pulse-laser break (Sup. Fig. 3I). This demonstrates that once Q-nMT bundles are formed, they do not need MT-kinetochore interaction to be maintained. Accordingly, Ndc80 or Ipl1 inactivation in cells that are already in quiescence had no effect on the Q-nMT bundle maintenance (Sup. Fig. 3B-C). Importantly, the above experiment indicates that once formed, Q-nMT stability is established and maintained throughout its length.

Specific kinesins are required for each step of Q-nMT bundle formation

To go further in the deciphering of the molecular mechanism involved in Q-nMT bundle formation, we focused on kinesins. Kar3 is a kinesin that, in complex with its regulator Cik1, can generate parallel MT bundles from a MT organizing center (MTOC) both in vitro and in proliferative cells (Manning et al., 1999; Mieck et al., 2015; Molodtsov et al., 2016). In quiescence, we found that Kar3-3GFP localized at the SPB, but also as dots along the Q-nMT bundle (Fig. 4A). In 4 day-old kar3Δ cells, the majority of the detected MT structures were extremely short compared to WT (Fig. 4B and Sup Fig. 4A). Same results were obtained in cik1Δ but not in vik1Δ cells, which lack the alternate Kar3 regulators (Fig. 4B). Thus, the Kar3/Cik1 complex is required for phase I.

Each phase of Q-nMT formation requires a specific kinesin.

(A) Images and corresponding line scans of WT cells (2 d) expressing Kar3-3GFP (green) and mTQZ-Tub1 (red).

(B) Morphometric Q-nMT bundle properties distribution in 4 d WT (grey), kar3Δ (red), vik1Δ (blue), cik1Δ cells (green) expressing mTQZ-Tub1 after Noc treatment. Blue crosses are SD. Each circle corresponds to an individual Q-nMT bundle. Representative cells are shown.

(C) Nuclear MT length distribution in WT and cin8Δ cells expressing mTQZ-Tub1 treated (dashed boxes) or not (plain boxes) with Noc.

(D) Fluorescence intensity along Q-nMT bundle in WT and cin8Δ cells expressing mTQZ-Tub1 7 h and 24 h after glucose exhaustion. Representative cells are shown.

(E) WT and cin8Δ cells expressing mEOS3.2-Tub1 were imaged using PALM (images are in pseudo-colors). Full width at half maximum (FWHM) was measured at the indicated distance from the SPB. Each line in the bottom graph corresponds to a single cell. P-value between WT and cin8Δ are indicated (unpaired T-test).

(F) WT cells (3 d) expressing Kip1-GFP (green) and mTQZ-Tub1 (red). Graphs show fluorescence intensity along normalized Q-nMT bundles (plain and dash lines: mean and SD respectively).

(G-H) Representative images and nuclear MT length distribution in WT, kip1Δ and kip3Δ cells expressing mTQZ-Tub1 treated (dashed boxes) or not (plain boxes) with Noc.

In C and H, *: p-value <0.05, and ***: p-value <10-6 (unpaired T-test). Means and SD are indicated. In A, B, D, E, F and G, bar is 2 µm.

Cin8 is a Kinesin-5 that cross-links MTs (Bodrug et al., 2020; Pandey et al., 2021; Singh et al., 2018), and as such, could play a role in Q-nMT bundle stabilization. Cin8 was hardly detectable in quiescent cells (Sup. Fig. 4B) and cin8Δ cells assembled Q-nMT bundles that were thinner than in WT cells (Fig. 4C-E). Yet, these thinner bundles became stabilized in phase III (Fig. 4C and Sup. Fig. 4C-D). Overall, these results strongly suggest that Cin8 is required for phase II MT nucleation and/or elongation but not for stabilization in phase III.

Yeast have an additional Kinesin-5 called Kip1 (Fridman et al., 2013). In quiescence, most of the Kip1-GFP signal was observed at the Q-nMT bundle +end (Fig. 4F). In kip1Δ cells, phase I was slightly delayed and Q-nMT bundle were not stabilized in phase III (Fig. 4G-H), yet they were as thick as WT Q-nMT bundles (Sup. Fig. 4E). These data demonstrate that Kip1 is required for Q-nMT bundle stabilization in phase III. Of note, in phase I and II, Q-nMT bundles were slightly longer in cells deleted for the Kinesin-8 Kip3, a MT depolymerase ((Fukuda et al., 2014) Fig. 4 G-H and Sup. Fig. 4G) but were unaffected in cells deleted for Kip2, a kinesin that stabilizes cytoplasmic MTs ((Hibbel et al., 2015) Sup. Fig. 4F-G). Taken together our results demonstrate that each phase of Q-nMT bundle assembly involves specific kinesins.

Q-nMT bundle disassembly conditions SPB duplication/separation

Finally, we questioned the molecular mechanism of Q-nMT bundle disassembly upon quiescence exit. Cycloheximide prevented Q-nMT bundle disassembly indicating that this process requires de novo protein synthesis (Fig. 5A). We then measured Q-nMT bundle thickness upon depolymerization, and found that while Q-nMT bundles shortened, they did not get thinner (Fig. 5B). In agreement, Nuf2-GFP clusters found at the Q-nMT bundle +end moved back to the SPB while the Nuf2-GFP clusters localized in the middle of the structure remained immobile until they are reached by these later (Sup. Fig. 5 A). This shows that all the MT +ends did not start depolymerizing at the same time, and that longer MTs depolymerized first. This finding is in agreement with our model in which MTs are cross-linked all along the Q-nMT bundle length.

Q-nMT bundle disassembly always occurs before SPB separation upon quiescence exit.

(A) WT cells expressing Spc42-RFP (red) and Nuf2-GFP (green) (5 d) were re-fed on an YPDA microscope pad. Individual Q-nMT bundles were measured in cells treated with CHX (blue, Stu2-GFP) or with DMSO alone (grey, Nuf2-GFP). Each line corresponds to an individual cell. For each cell, time was set to zero at the onset of MT bundle depolymerization (black dashed line).

(B) Q-nMT bundle length (green) and fluorescence intensity at full width half maximum (FWHM - orange) were measured upon quiescence exit in WT cells (5 d) expressing mTQZ-Tub1. Representative example of shirking Q-nMT bundle is shown on the left.

(C) Cells of the indicated genotype expressing mTQZ-Tub1 were grown for 4 d, and re-fed. Q-nMT bundle length was measured at the indicated time points, 15 min after a Noc treatment.

(D) Western blot (GFP antibodies) on total protein extracts from WT cells expressing Kip3-GFP grown for the indicated time. Sac6 was used as a loading control (Sac6 antibodies).

(E) WT and kip3Δ cells (5 d) expressing Nuf2-GFP were re-fed on an YPDA microscope pad. Each line corresponds to an individual cell. For each cell, time was set to zero at the onset of SPB separation.

(F) Representative images of kip3Δ cells expressing Nuf2-GFP upon quiescence exit. White arrowheads: Q-nMT bundle extremities.

(G) Percentage of cells expressing Spc42-mRFP1 with separated SPB as a function of time upon quiescence exit.

(H) WT and kip3Δshe1Δ cells (6 d) expressing Spc42-mRFP1 (red) and mTQZ-Tub1 (green) were re-fed on a YPDA microscope pad. Percentage of cells with a single or a duplicated SPB in the presence or absence of Q-nMT bundle were scored. Representative cells are shown. Right bottom panel: actin (phalloidin staining, red) in kip3Δ she1Δ cells (6 d) expressing mTQZ-Tub1 (green) before and 1 h after quiescence exit.

In F and H, blue arrowheads: SPB. In all panels, bar is 2 µm.

Since Kip3 is the only well-characterized yeast MT depolymerase, we studied Q-nMT bundle disassembly in kip3Δ cells and observed that it was drastically delayed (Fig. 5C), yet its overall depolymerization rate remained unaffected (Sup. Fig. 5B). In fact, we found that in WT cells, Kip3 needs to be resynthesized upon quiescence exit (Fig. 5D). We searched for proteins that could be involved in Q-nMT bundle disassembly together with Kip3. Among proteins required for mitotic spindle disassembly, we found that inactivation of CDC14, CDC15, CDH1, DCC1 and TOR1 had no effect on Q-nMT bundle disassembly, just as with the inactivation of Ipl1 or proteins involved in kinetochore-MT attachment (Sup. Fig. 5 C to F). The sole additional actor we found was She1, a dynein regulator (Bergman et al., 2012), which deletion strongly aggravated the kip3Δ phenotype (Fig. 5C). Thus, Q-nMT bundle disassembly does not rely on the canonical mitotic spindle disassembly machinery but, rather, specifically required Kip3 and She1.

Importantly, when we followed quiescence exit at the individual cell level, Q-nMT bundle depolymerization always preceded SPB duplication/separation (Fig. 5A) and thus, in both WT and in kip3Δ cells (Fig. 5E, F). In fact, when we impeded Q-nMT bundle disassembly using kip3Δ she1Δ, we found that SPB duplication was severely delayed (Fig. 5G-H, Sup. Fig. 5G-H). Importantly, kip3Δ she1Δ cells proliferated just like WT cells (Sup. Fig. 5G) and upon quiescence exit, these mutant cells disassembled actin bodies just as in WT cells (Sup. Fig. 5I), demonstrating that kip3Δ she1Δ cells were not impaired in sensing quiescent exit signals. These observations indicate that Q-nMT bundle has to be disassembled to allow SPB duplication/separation upon quiescence exit and cell cycle re-entry.

Discussion

Our results shed light on the precise temporal sequences that generate a Q-nMT bundle upon quiescence establishment in yeast. From bacteria to human stem cells, modifications of the cell physico-chemical properties are known to accompany quiescence establishment. Among these modifications, a reduction of the cellular volume increases molecular crowding (Joyner et al., 2016) and pH acidification causes changes in the surface charges of macromolecules and viscosity increases (Charruyer and Ghadially, 2018; Jacquel et al., 2021; Munder et al., 2016; Persson et al., 2020). Much evidence points to these modifications as a trigger for the auto-assembly of several types of enzyme-containing granules (Munder et al., 2016; Petrovska et al., 2014; Rabouille and Alberti, 2017), or complex structures such as P-bodies and Proteasome Storage Granules (Peters et al., 2013; Jacquel et al., 2021; Currie et al., 2023). Recently, Molines and colleagues showed that cytoplasmic viscosity modulates MT dynamics in vivo (Molines et al., 2022). Here, we show that Q-nMT bundle formation does not rely on modifications of the physicochemical properties that cells experience upon quiescence establishment (Sup. Fig. 1). In fact, dynamic cMTs and a stable Q-nMT bundle can be observed simultaneously in a quiescent cell (Sup. Fig. 1H), while physicochemical properties evolve in parallel both in the nucleus and in the cytoplasm (Joyner et al., 2016). Importantly, reduction of the cellular volume, pH acidification and increased viscosity are observed within minutes upon glucose starvation (Joyner et al., 2016) while the Q-nMT bundle needs a couple of days to be fully assembled (Fig. 1A). Further, we show that its formation cannot be delayed in time (Fig. 1H) and as such probably depends on an active process induced by a transient signal emitted upon glucose exhaustion. We speculate that the SPB may act as a platform that integrates this signal, and transfers it to the MT machinery in order to assemble a stable parallel structure.

As it is required for the onset of Q-nMT bundle formation (Fig. 3B), Aurora B/Ilp1, a kinase that plays a key role at the MT-kinetochore interface where it senses tension, could be one of the target of the above-mentioned nutritional signal. In parallel, upon quiescence establishment, chromosomes hyper-condensation (Guidi et al., 2015; Rutledge et al., 2015) may modify tension at the kinetochore/MT interface and as such could contribute to the initiation of phase I. Besides, Bim1 has also been involved in MT-kinetochore attachment (Dudziak et al., 2021; Akhmanova and Steinmetz, 2015). Although it deletion alone has no effect on Q-nMT bundle formation, its role become critical for phase I if kinetochore-MT interaction are already destabilized by the absence of other CPC components such as Sli15 and Bir1 (Fig. 3D and Sup. Fig. 3E). Altogether, these observations point to kinetochore-MT interaction as essential for phase I initiation, as confirmed by the absence of Q-nMT bundle in the ndc80-1 mutant (Fig. 3A).

In phase I, SPB-anchored MTs elongate and are concomitantly stabilized (Fig. 1 A-C). This step requires the kinesin-14 Kar3 and its regulator Cik1 (Fig. 4B). During mating (Molodtsov et al., 2016), and in early mitosis, when half-spindles form (Kornakov et al., 2020), Kar3/Cik1, together with Bim1, align and cross-link growing MTs along existing ones, thereby promoting the organization of MTs into parallel bundles. Yet, in mitosis, the kinesin-14/EB1 complex promotes MT dynamics (Kornakov et al., 2020). In phase I, whether the robust MT stabilization depends on the modification of the kinesin-14/EB1 complex properties or is due to additional specific cross-linker(s) remains to be clarified (Fig. 6b).

Model for Q-nMT bundle assembly

MT-kinetochore interaction and Ilp1 (Aurora B) are required for the onset of phase I. Kar3 (kinesin-14) and its regulator Cik1 are essential to initiate Q-nMT bundle elongation. Although deletion of BIM1 (EB1) has no effect, it becomes critical for phase I if kinetochore-MT interactions are already destabilized by the absence of Chromosome Passenger Complex components. Kinetochore clustering by the monopolin complex and Slk19 is needed to maintain MT bundling while phase I MTs elongate. During phase I, Tub4 (ɣ-Tubulin) accumulates at the SPB. In phase II, a second wave of MT nucleation and elongation occur. Phase II MTs are concurrently stabilized along pre-existing phase I MTs, in a Cin8 (kinesin-5)-dependent manner. Phase I and phase II MTs + end (>1 µm) remain dynamic until the full length Q-nMT bundle stabilization is reached via the action of Kip1 (kinesin-5), about 2 days after glucose exhaustion.

In quiescence, deletion of either monopolin or Slk19 leads to the formation of short, shattered and flared MT structures, phenotypes that are worsened by the deletion of Bim1 (Fig. 3F-H and Sup. Fig. 3F). Since Monopolin and Slk19 are involved in kinetochore clustering (Plowman et al., 2019; Rabitsch et al., 2003; Tóth et al., 2000; Lee et al., 2004; Mittal et al., 2019; Movshovich et al., 2008; Richmond et al., 2013; Zeng et al., 1999), we speculate that kinetochore-kinetochore interaction may not only help preventing depolymerization of single MT, but also constrain elongating phase I MTs and facilitate their concomitant cross-linking (Fig. 6c).

With the initiation of phase I, Tub4 starts accumulating at the SPB to allow the second wave of MT nucleation observed in phase II. When phase II MTs elongate, they are concurrently stabilized along pre-existing phase I MTs (Fig. 6d). This step relies on the kinesin-5/Cin8 as Q-nMT bundles assembled in cin8Δ cells are thinner than WT Q-nMT bundles (Fig. 4C and D). Intriguingly, in vitro, the monomeric form of the vertebrate kinesin-5 Eg5 promotes MT nucleation and stabilizes lateral tubulin-tubulin contacts (Chen et al 2019). We thus assume that in the absence of Cin8, either nucleation of phase II MTs cannot occur, or phase II MTs can elongate, but cannot be cross-linked and stabilized to phase I MT while they grow, and as such rapidly vanish. An absence of bundle thickening in phase II is also observed in the “Tub1-only” mutant i.e. in the absence of Tub3. Tubulin isoforms or tubulin PTM can allow the recruitment of specific kinesins and thereby modify MT properties, see for example (Sirajuddin et al., 2014; Peris et al., 2009; Dunn et al., 2008). May Tub3 be important for Cin8 recruitment or function during phase II?

Then, MTs elongate, but their distal parts (>1 µm) are not yet stable (Fig. 1A). The full-length Q-nMT bundle stabilization is reached about 2 days after glucose exhaustion. This step depends on the kinesin-5/Kip1 (Fig. 4 G-H and Sup. Fig. 4E) that is essential to cross-link and stabilize elongating MTs during phase III (Fig. 6e). Accordingly, we observed a Kip1 enrichment at the Q-nMT bundle +ends (Fig. 4F). Thus, in agreement with their ability to form homo-tetramers able to cross-link (Acar et al., 2013; Kapitein et al., 2005; Scholey et al., 2014; Weinger et al., 2011) and stabilize parallel MTs (Kapitein et al., 2005; Shimamoto et al., 2015; Yukawa et al., 2020), the two yeast kinesin 5 are essential for Q-nMT bundle formation. Yet, in quiescence, just as in mitosis, Kip1 and Cin8 have clearly non-equivalent functions (Roostalu et al., 2011; Shapira and Gheber, 2016).

So far, our results indicate that the Q-nMT bundle formation mechanism is different from the one at work during mitotic spindle assembly. Similarly, we show that Q-nMT bundle disassembly relies on a mechanism that does not involves the same pathway than mitotic spindle dismantlement. Indeed, Q-nMT bundle disassembly involves Kip3 (Fig. 5C-D) but is independent of the anaphase-promoting complex and Aurora B (Woodruff et al., 2010). Interestingly, the longest MTs start depolymerizing from their +ends first. Then, when they reach the +ends of shorter MTs, they embark them on depolymerization (Sup. Fig. 5A). This cooperative behavior has been observed in vitro within parallel MT bundles (Laan et al., 2008) and our data indicate that such a comportment can exist in vivo. In telophase, the CPC localizes to the end of the two half-spindles in order to participate in mitotic spindle disassembly (Ibarlucea-Benitez et al., 2018). Intriguingly, in quiescence, CPC is poised at the Q-nMT bundle + end (Fig. 3C). May this CPC localization facilitate Q-nMT bundle rapid disassembly once environmental conditions become favorable, thus ensuring quiescence exit efficiency?

Finally, in the present study, we show that the Q-nMT bundle is required for WT cells survival in quiescence. While we do not yet know why this structure is important to face chronological age, it is clear that the presence of Q-nMT bundle strongly alters the nuclear organization (Laporte et al., 2013; Laporte and Sagot, 2014; Laporte et al., 2016). As chromatin organization influences gene expression, one hypothesis could be that the presence of Q-nMT bundle is required for expression of genes required for cell survival in quiescence. Another attractive hypothesis could be that the Q-nMT bundle is the yeast analog of the primary cilium of mammalian cells. In fact, these two structures formed upon proliferation cessation, they are both composed of highly stable parallel MTs, their formation involve MT motors, and they are both template from the centriole/SPB. More importantly, we show here that the Q-nMT disassembly is needed to authorize nascent SPB separation upon quiescence exit (Fig. 5). Thus, the Q-nMT bundle disassembly may controls reentry into the proliferation cycle, just as the primary cilium does in ciliated mammalian cells (Goto et al., 2017; Kim and Tsiokas, 2011).

Materiel and Methods

Yeast strains, plasmids and growth conditions

All the strains used in this study are isogenic to BY4741 (mat a, leu2Δ0, his3Δ0, ura3Δ0, met15Δ0) or BY4742 (mat alpha, leu2Δ0, his3Δ0, ura3Δ0, lys2Δ0), available from GE Healthcare Dharmacon Inc. (UK), except in Fig. 2, in which S288c strains were used (Nsamba et al., 2021). BY strains carrying GFP fusions were obtained from ThermoFisher Scientific (Waltham, MA, USA). Integrative plasmid pTub1-mTurquoise2-Tub1, pTub1-wasabi-Tub1, pTub1-mRUBY2-Tub1, pTub1-mEOS2-Tub1 were a generous gift of Wei-Lih Lee (Markus et al., 2015). The strain expressing SPC42-mRFP1 is a generous gift from E. O’Shea (Huh et al., 2003). The strain expressing TUB4-mTQZ is generous gift of S. Jaspersen (Burns et al., 2015). Three copies of eGFP in tandem were integrated at the 3’ end of DAD2, TUB3 or KAR3 endogenous loci respectively. Plasmids for expressing Nup2-RFP (p3695) or Bim1-3xGFP (p4587) from the endogenous locus were described in (Laporte et al., 2013).

For Fig. 1F, pHIS3:mTurquoise2-Tub1+3’UTR::LEU2 was integrated at TUB1 locus, then a pRS303-ADH2p-mRuby2-Tub1 was integrated at the HIS3 locus. To generate this plasmid, the ADH2 promoter was amplified from yeast genomic DNA flanked with NotI and SpeI restrictions sites and inserted in pRS303. The mRuby2-Tub1+3’UTR, including the 116-nucleotide intron and 618 nucleotides downstream the stop codon was cloned between SpeI/SalI sites.

Yeast cells were grown in liquid YPDA medium at 30 °C in flasks as described previously in (Sagot et al., 2006) except for experiments using thermosensitive strains, where cells were first grown at 25 °C then shifted at 37 °C for indicated time before imaging.

Experiment in Sup. Fig. 1I was performed as described in (Orij et al., 2009, Peters et al., 2013). In brief, proliferating cells were transfered at an OD600nm of 0.5 in Hepes buffer (25 mM Hepes, pH 7.4, 200 mM KCl, 1 mM CaCl2, and 2% dextrose) buffered either at pH 4 or 7.5, in the presence of 100 µM CCCP (Sigma-Aldrich). After 150 min 30 °C shaking, cells were imaged (for mTQZ-Tub1) or fixed and stained using Alexa Fluor 568–phalloidin (Invitrogen) as described (Sagot et al., 2006).

For live cell imaging, 2 μL of the cell culture were spotted onto a glass slide and immediately imaged at room temperature.

For quiescence exit, cells were then incubated 2-3 min in liquid YPD and a 2 µL were spread onto a 2 % agarose microscope pad containing YPD. Individual cells were imaged every hour, up to 6 or 12 h at 21°C. For quiescence exit in presence of cycloheximide (Fig. 5A), cells were pre-incubated 30 min in the presence of the drug prior quiescence exit. Cycloheximide was used at 180 µM (Sigma-Aldrich), Nocodazole was used at 30 µg/mL (7.5 µM) (Sigma-Aldrich).

Fluorescence Microscopy

Cells were observed on a fully-automated Zeiss 200M inverted microscope (Carl Zeiss, Thornwood, NY, USA) equipped with an MS-2000 stage (Applied Scientific Instrumentation, Eugene, OR, USA), a Lambda LS 300 Watt xenon light source (Sutter, Novato, CA, USA), a 100X 1.4NA Plan-Apochromat objective, and a 5 position filter turret. For RFP imaging we used a Cy3 filter (Ex: HQ535/50 nm – Em: HQ610/75 nm – BS: Q565 nm lp). For GFP imaging, we used a FITC filter (excitation, HQ487/25 nm; emission, HQ535/40 nm; beam splitter, Q505 nm lp). For mTurquoise imaging we used a DAPI filter (Ex: 360/40 nm – Em: 460/50 nm– BS: 400 nm). All the filters are from Chroma Technology Corp. Images were acquired using a CoolSnap HQ camera (Roper Scientific, Tucson, AZ, USA). The microscope, camera, and shutters (Uniblitz, Rochester, NY, USA) were controlled by SlideBook software 5. 0. (Intelligent Imaging Innovations, Denver, CO, USA).

For PALM (Fig. 4E) a Nikon Ti-Eclipse equipped with iLas2 TIRF arm, laser diodes (405, 488, 532, 561, 642 nm), a 100x 1.49 oil (TIRF) objective connected to a EMCCD Camera Photometrics Evolve was used.

Immunofluorescence was done as in (Laporte et al. 2015), using AlexaFluor Phalloidin (Invitrogen).

For expansion microscopy (Sup Fig. 2A), spheroplasts were obtained as described in (Laporte et al., 2016). After washes in 10 mg/mL NaBH4 in PEMS for 10 min, spheroplasts were seeded on 12 mm round poly-L-lysine coated coverslips, washed twice for 30 min in 100 μL PEM-BAL (PEM + 1% BSA) and incubated with the primary antibodies (anti-GFP from mouse; Roche, ref. 11814460001, 1/50; anti-alpha-tubulin from rat; YOL1/34, Abcam, ref. Ab6161, 1/100) for 1 hour at 37°C. Cells were then washed three times in PEM-BAL and incubated with the secondary antibodies (Goat anti-mouse AlexaFluor®488 1/200 and Donkey anti-rat AlexaFluor®555; A11029 and A21434 respectively, ThermoFisher Scientific, Waltham, MA) for 45 min at 37 °C. Cells were then washed three times with PBS. Cells were processed for Expansion Microscopy as previously described in (Bahri et al., 2021). Spheroplasts were incubated for 10 min in 0.25 % GA in PBS, washed in PBS three times for 5 min, and processed for gelation. A drop of 100 µL of ExM MS (8.625% (wt/wt) SA, 2.5% (wt/wt) AA, 0.15 % (wt/wt) BIS, 2 M NaCl in 1× PBS) was placed on the chilled Parafilm, and coverslips were put on the drop with cells facing the solution and incubated for 3 min. Then the coverslips were transferred to a 35 μL of ExM MS supplemented with 0.2 % APS and 0.2 % TEMED, with the initiator (APS) added last. Gelation proceeded for 3 min on ice, and samples were incubated at 37 °C in a humidified chamber in the dark for 1 h. Then coverslips with attached gels were transferred into a six-well plate for incubation in 2 mL of digestion buffer (1× TAE buffer, 0.5% Triton X-100, 0.8 M guanidine hydrochloride, pH ∼8.3) supplemented with DAPI (1 µg/mL) for 10-15 min at 37 °C, until the gels detached. Fresh proteinase K at 8 units/mL was then added and samples incubated at 37 °C for 30 min. Finally, gels were removed and placed in 10mL petri dishes filled with ddH2O for expansion. Water was exchanged at least twice every 30 min, and incubated in ddH2O overnight at RT. Gels expanded between 4× and 4.2× according to SA purity. Expanded cells were imaged with a UPlanS Apo 100×/1.4 oil immersion objective in an Olympus IX81 microscope (Olympus, Tokyo, Japan). For Structured illumination microscopy (SIM, Fig. 7D), a ZEISS Lattice Lightsheet 7 was used.

Laser ablation (Fig. 3I) was performed at room temperature with a 100X oil Plan-Apochromat objective lens (NA 1.4) and an Axio-Observed.Z1 microscope (Carl Zeiss) equipped with a spinning disk confocal (Yokogawa), an EMCCD Evolve camera (Photometrics and Roper Scientific) and 491 nm (100 mW; Cobolt calypso) and 561 nm (100 mW; Cobolt Jive) lasers. Images were acquired with Metamorph software (Roper Scientific). Every 5 to 20 s, a Z-series of 0.4 µm steps were acquired. A 355 nm microchip laser (Teem Photonics) with a 21 kHz repetition rate, 0.8 µJ energy/pulse, 2 kW of peak power and 400 ps pulse width, powered with an iLas2 PULSE system (Roper Scientific) was used between 10 to 40 % power with one pulse of a spot length of 100 points. Breakage was considered as successful if a non-alignment between the two remaining Q-n MT bundle fragments was observed.

Z-stacks were deconvolved using the Deconvolution Lab plugin (Fig. 1D; Fig. 3C,E,F Fig. 4A,F). In figures 1F, 4A, B, G, supplemental figure 1H and 3B a faint and fuzzy fluorescence signal is detected in some cell cytoplasm using the GFP filter set. This signal is not GFP but rather due to a non-specific yellow background fluorescence.

Image analysis

Distribution and associated statistics were performed using GraphPad Prism 5 (GraphPad Software, Inc. La Jolla, USA). Unless specified, histograms and scatter dot plots show the mean and the error bars indicate SD. MT length was measured on MAX-projected image using imageJ.

For MT fluorescence intensity measurement, a line scan (i1) of 4 pixel width containing both FP signal and background was drawn along MTs on SUM projection image (3-4 Z-plans) using ImageJ. A line of 8 pixels width (at the same location) was drawn in order to calculate the intensity of the surrounding background (i2). The real intensity (ir) was calculated as follow: int=(i1-((i2×8)-(i1×4))/4). Similar approach was conducted to measure Tub4 fluorescence intensity (Fig. 1E) i.e. (i1) and (i2) were two boxes, the second including the first one. Intensity was calculated as follow: int=[i1-([(i2xi2surface)-(i1xi1surface)]/(i2surface-i1surface))*i1surface]. For morphometric Q-nMT bundle property distribution (Fig. 1C and 4B)the mean fluorescence intensity was measured in each individual cell, in a finite area localized adjacent to the SPB (0,4 µm – 3-4 Z stack sum projected) as an estimate of Q-nMT bundle width and plotted as a function of the according measured bundle length.

For Nuf2 fluorescence intensity measurement at SPB (Sup. Fig. 1D), GFP and RFP line scan measurements were done on a 3-4 Z-plans sum projection. The “SPB localization zone” was defined as the length ending at the second pixel after the brightest RFP signal of Spc42-RFP. The remaining Q-nMT bundle zone was defined as the “+end zone”. The total Nuf2-GFP signal detected along the Q-nMT bundle was set to 100 % (sum projection imaged) allowing to determine the percentage of the signal measured at “SPB localization zone” and at “+end zone”

Normalized Q-nMT bundle length (Fig. 3C) was done after line scan intensity measurement. mTQZ-Tub1 intensities slopes were first aligned on their inflexion point. To compare Q-nMT bundle with different lengths, we first sorted Q-nMT bundle <1.8 µm. After an artificial isotropic expansion, we fit all MT structures to length order. Then, corresponding mean intensity of Q-nMT bundle (mTZQ-Tub1) and GFP signal were calculated.

To measure the Q-nMT bundle depolymerization (Fig. 5A and E), individual Q-nMT bundle length were measured over time. The first measurement was set at 100 % of the length, and the following percentages calculated accordingly. A fluorescence drop above or equal to 25 % defines the inflexion point of the slope, and was used to align the different length measurements. The FWHM (Full Width at Half Maximum) was calculated by measuring fluorescence intensity of a line crossing perpendicularly a Q-nMT bundle, from a sum projection. After fitting intensity level with Gaussian distribution and obtained the standard deviation (σ) value, FWHM was calculated using the equation 2√(2ln2)xσ. In Fig. 5B, FWHM was measured at 0.5 µm of the SPB.

Western blots

Western blots were done as described in (Sagot et al., 2006) using anti-GFP antibodies (Roche); anti-Tat1 antibodies (a generous gift from J-P. Javerzat) and antibody against the budding yeast Act1 or Sac6, generous gifts from B. Goode.

Phenotypical analysis of cells without Q-nMT bundle

WT strains were grown to glucose exhaustion (OD ∼6.5), washout with “old YPDA” and pulsed with either DMSO or Nocodazole (30µg/mL) for 24 h. Cells were then washed twice in “old YPDA” before being incubated in the same medium. Dead cell measurement was done using methylene blue staining. The capacity of cells to exit quiescence was scored after cells micro-manipulation (N=4, n>100) as described in (Laporte et al., 2011).

Supporting information

eLife Assessment

This work presents important insights regarding the mechanism underlying the assembly, maintenance, and disassembly of a very stable microtubule-based structure, termed quiescent-cell nuclear microtubule (Q-nMT) bundle, formed in quiescent yeast cells to ensure cell survival and viability. Some of the evidence is solid, but some of the major claims are only incompletely supported and require additional analyses using state-of-the-art methodology and more precise descriptions.

Reviewer #1 (Public Review):

In their manuscript, Laporte et al. analyze the process of formation of the quiescent-cell nuclear microtubule (Q-nMT) bundle, a set of parallel MTs that emanate from the nuclear side of the spindle pole bodies (SPBs) upon the entry of Saccharomyces cerevisiae cells in quiescence. Based on their results, the authors propose that Q-nMT bundle formation is a multistep process that comprises three distinct sequential phases. The authors further evaluate the role of different factors during the growth of the Q-nMT bundle upon quiescence entry, as well as during the disassembly of this structure once cells resume their proliferation.

The Q-nMT is an interesting cellular structure whose physiological function is still widely unknown. Hence, providing new insights into the dynamics of Q-nMT bundle formation and identifying new factors involved in this process is an interesting topic of relevance in the field. The authors made a substantial effort in order to evaluate Q-nMT bundle formation and provide a considerable amount of data, mainly obtained from microscopy analyses. Overall, the experiments are well described and properly executed, and the data in the manuscript are clearly presented.

Despite the interest in the study, there are important issues that could affect the validity of the conclusions drawn in the manuscript. In this way, regarding the analysis of the dynamics of Q-nMT bundle formation, the experimental set up described in some of the experiments raises certain concerns, which mostly derive from the nocodazole treatments and the use of this microtubule-depolymerizing agent as the only approach to evaluate the stability of the Q-nMT bundle. On the other hand, regarding the factors involved in Q-nMT formation, the differences in microtubule length with the wild-type strain, despite being statistically significant, are really subtle for many of the mutants analyzed (e.g., bir1, slk19, etc.). Additionally, there are proteins that are proposed to participate in the process of Q-nMT formation and whose expression during quiescence needs to be demonstrated. Finally, although the cell viability defects observed for some of the mutants in these factors could be certainly associated with the lack of expression or mutation of the specific gene under evaluation, in none of the cases can they be directly attributed to a defect in aberrant Q-nMT bundle formation.

Based on the aforementioned reasons, and despite the considerable effort by the authors, it is my impression that many of the conclusions of the manuscript are not sufficiently justified by the data provided. Additional evidence, including the incorporation of key experimental controls that are currently missing, would be required in order to more strongly support the conclusions of the manuscript.

Reviewer #2 (Public Review):

Summary: The authors investigate the assembly of the Q-nMT, a stable microtubule structure that is assembled during quiescence. Notably, the authors show that the formation of the Q-nMT cannot be solely explained by changes in the physicochemical properties of quiescent cells. The authors report that Q-nMT assembly occurs in three regulated steps and identify kinesin motor proteins involved in the assembly and disassembly of the structure.

Strengths: The findings provide new insight into the assembly and possible function of the Q-nMT with respect to the response of haploid budding yeast to glucose starvation.

Weaknesses: The manuscript would benefit from more precise language and requires additional clarification regarding how claims are supported by the evidence. Clear definitions are also required, for example, "active process" is not defined. Some conclusions are not supported by the results, for example, the claim that the Q-nMT functions as a checkpoint effector that inhibits re-entry into the cell cycle.

Reviewer #3 (Public Review):

In this study, the authors analyzed a unique and very stable microtubule bundle that is formed in yeast cells entering quiescence. They show that the structure is required for yeast cells to maintain viability during quiescence and that it needs to be disassembled for cell cycle re-entry. They identify different stages during the assembly process and focus on the molecular players required for microtubule bundle formation and stabilization. They identify kinetochore as well as molecular motors such as auroraB, kinesin-14, and kinesin-5 that assemble, stabilize and crosslink the microtubules of the bundle. The paper also investigates the disassembly of the structure and shows that disassembly is required for cell cycle re-entry.

The study is very comprehensive, provides quantifications to support claims, and identifies various players involved in these processes, providing mechanistic insight. It also presents various control experiments to exclude alternative explanations and support the proposed model.

It is the first molecular-level insight into how this very stable microtubule structure can be assembled, maintained, and disassembled, and how this is coordinated with cell cycle exit and re-entry. This information may be very useful when analyzing similarly stable, microtubule-based structures in other organisms such as cilia in animals, which also display cell cycle-coordinated dynamics.

Overall, this is a nicely presented study that provides important insight into the field and beyond, but there are a few points that need to be addressed regarding methods used for quantifications and data presentation.

Author Response

The following is the authors’ response to the original reviews.

Reviewer #1 (Recommendations for The Authors):

To hopefully contribute to more strongly support the conclusions of the manuscript, I am including a series of concerns regarding the experiments, as well as some recommendations that could be followed to address these issues:

(1) The Q-nMT bundle is largely unaffected by the nocodazole treatment in most phases during its formation. However, cells were only treated with nocodazole for a very short period of time (15 min). Have the authors analyzed Q-nMT stability after longer nocodazole exposures? Is a similar treatment enough to depolymerize the mitotic spindle? This result could be further substantiated by treatment with other MT-depolymerizing agents. Furthermore, the dynamicity of the Q-nMT bundle could be ideally also assessed by other techniques, such as FRAP.

The experiments suggested by the reviewer have been published in our previous paper (Laporte et al, JCB 2013). In this previous study, we presented data demonstrating the resistance of the Q-nMT bundle to several MT poisons: TBZ, benomyl, MBC (Sup Fig 2D) and to an increasing amount of nocodazole after a 90 min treatment (Sup Fig2E). These published figures are provided below.

The nMT array contains highly stable MTS.

(A) Variation Of nuclear MT length in function Of time (second) in proliferating cells. Cells express GFP•Tubl (green) and Nup2•RFP (red). Bars, 2 pm. N = l, n is indicated. (B) Variation of the nMT array length in function of time measured for BirnlGFP—expressing cells In = 161, for 6-d•old Dad2GFP—expressing cells In = 171, for Stu2GFP—expressing cells (n = 17), and 6•d-old Nuf2• GFP—expressing cells (n = 17). Examples Of corresponding time lapse are shown. Time is in minutes experiments). Bar, 2 pm. (CJ Nuf2•GFP dots detected along nMT array (arrow) are immobile. Several time lapse images of cells are shown. Time is in minutes. gar, 2 pm _ MT organizations in proliferating cells and 4-d•old quiescent cells before and after a 90-min treatment With indicated drugs. Bar, 2 pm. (E) MT organizations in Sci-old quiescent cells before and after a 90min treatment With increasing concentrations Of nocodazole.

In the same article, we showed that Q-nMT bundles resist a 3h nocodazole treatment, while all MT structures assembled in proliferating cells, including mitotic spindle, vanished (see Fig 2E below). In addition, in our previous article, FRAP experiments were provided in Fig 2D.

The nuclear array is composed of stable MTS.

Variation of the length in function of time of (A) aMTs in proliferating cells, (B) nMT array in quiescent cells (7 d), and the two MT structures in early quiescent cells (4 d). White arrows point ot dynamic aMTs. In A—C, N = 2, n is indicated ID) FRAP on 7-d-old quiescent cells. White arrows point to bleach areas. Error bars are SEM. In A—D. time is in seconds. (E) nMT array is not affected by nocodazole treatment. Before and various times after carbon exhaustion (red dashed line), cells were incubated for 3 h with 22.5 pg/pL nocodozole and then imaged. The corresponding control experiment is shown in Fig I A. In all panels, cells expressing GFP-TtJbl (green) and Nup2-RFP (red) are shown; bars, 2 pm.

This previous study was mentioned in the introduction and is now re-cited at the beginning of the results section (line 107-108).

As expected from our previous study, when proliferating cells were treated with Noc (30 µg/ml) in the same conditions as in Fig1, most of the short and the long mitotic spindles vanished after a 15 min treatment as shown in the graph below.

Proliferating cells expressing NOf2=GFP and mTQZ-TUb1 (00—2) were treated or not With NOC (30vgfmI) for 15 min.

% Of cells With detectable MT and representative cells are shown. Khi-teet values are indicated. Bar: 2 pm,

(2) The graph in Figure 1B is somewhat confusing. Is the X-axis really displaying the length of the MTs as stated in the legend? If so, one would expect to see a displacement of the average MT length of the population as cells progress from phase II to phase III, as previously demonstrated in Figure 1A. Likewise, no data points would be anticipated for those phases in which the MT length is 0 or close to 0. Moreover, when the length of half pre-anaphase mitotic spindle was measured as a control, how can one get MT lengths that are equal or close to 0 in these cells? The length of the pre-anaphase spindle is between 2-4 um, so MT length values should range from 1 to 2 um if half the spindle is measured.

The graph in Fig1B represents the fluorescence intensity (a proxy for the Q-nMT bundle thickness) along the Q-nMT bundle length.

Fluorescence intensity is measured along a “virtual line” that starts 0,5 µm before the extremity of the QnMT bundle that is in contact with the SPB. In other words, we aligned all intensity measurements at the fluorescence increasing onset on the SPB side. We arbitrarily set the ‘zero’ at 0,5um before the fluorescence increased onset. That is why the fluorescence intensity is zero between 0 and 0,5 µm – The X-axis represents this virtual line, the 0 being set 0,5 µm before the Q-nMT bundle extremity on the SPB side. This virtual line allows us to standardize our “thickness” measurements for all Q-nMT bundles.

Using this standardization, it is clear that the length of the Q-nMT bundles increased from phase II to III (see the red arrow). Yet, as in phase II, Q-nMT bundles are not yet stable, their lengths are shorter in phase II than in phase II after a Noc treatment (compare the end of the orange line and the end of the blue line in phase II).

This is now explained in details in the Material and Methods section (line 539-545).

This is the same for the inset of Fig 1B and in Sup Fig 1A, in which we measured fluorescence intensity along the halfmitotic spindle just as we did for MT bundle. The X-axis represent a virtual line along the mitotic spindle, starting 0,5 µm before the SBP spindle extremity.

(3) Microtubules seem to locate next to or to extend beyond the nucleus in the control cells (DMSO) in Figure 1H. Since both nuclear MTs and cytoplasmic MTs emanate from the SPBs, it would have been desirable to display the morphology of the nucleus when possible. Moreover, since the nucleus is a tridimensional structure, it would also be advisable to image different Z-sections.

Analysis demonstrating that Q-nMT bundles are located inside the nucleus have been provided in our previous paper (Laporte et al, JCB 2013). In this article most of the images are maximal projections of Z-stacks in which the nuclear envelope is visualized via Nup2-RFP (see Fig1 of Laporte et al, JCB 2013 as an example below).

MTsare organized as a nuclear array in quiescent cells.

(A) MT reorganization upon quiescence entry. Cells expressing GFP-Tub1 (green) and Nup2RFP (red) are shown. Glucose exhaustion is indicated as a red dashed line. Quiescent cells dl expressing Tub I-RFP and either Spc72GFP,

In Laporte et al, JCB 2013, we also provided EM analysis both in cryo and immune-gold (Fig 1E below).

(top) or coexpr;sse8 with Tub I-RFP (bottom).

Arrows point dot along the nMT array. Bars: (A—C)) 2 pm. (E) AMT arroy visualized in WT cells by EMI Yellow arrows, MTS; red arrowheads, nuclear membrane; pink arrow, SPB. Insets: nMT cut transversally. Bar, 100 nm.

(4) Movies depicting the process of Q-nMT bundle formation in live cells would have been really informative to more precisely evaluate the MT dynamics. Likewise, together with still images (Fig 1D and Supp. Fig. 1D), movies depicting the changes in the localization of Nuf2-GFP would have further facilitated the analysis of this process.

In a new Sup Fig 1E, we now provide images of Q-nMT bundle formation initiation in phase I, in which it can be observed that Nuf2-GFP accompanies the growth of MT (mTQZ-TUB1) at the onset of Q-nMT bundle formation. Unfortunately, it is technically very challenging to follow the entire process of Q-nMT bundle formation in individual cells, as it takes > 48h. Indeed, for movies longer than 24h, on both microscope pads or specific microfluidic devices (Jacquel, et al, eLife 2021), phototoxicity and oxygen availability become problematic and affect cells’ viability.

(5) Western blot images displaying the relative protein levels for mTQZ-Tub1 and of the ADH2 promoter-driven mRuby-Tub1 at the different time points should be included to more strongly support the conclusion that new tubulin molecules are introduced in the Q-nMT bundle only after phase I. It is worth noting, in this sense, that the percentage of cells with 2 colors Q-nMT bundle is analyzed only 1 hour after expression of mRuby-Tub1 was induced for phase I cells, but after 24 hours for phase II cells.

We have modified Fig 1F and now provide images of cells after 3, 6 and 24h after glucose exhaustion and the corresponding percentage of cells displaying Q-nMT bundle with the two colors. We also now provide a western blot in Sup Fig 1H using specific antibodies against mTQZ (anti-GFP) and mRuby (anti-RFP).

(6) In order to demonstrate that Q-nMT formation is an active process induced by a transient signal and that the Q-nMT bundle is required for cell survival, the authors treated cells with nocodazole for 24 h (Fig 1H and Supp Fig 1K). Both events, however, could be associated with the toxic effects of the extremely prolonged nocodazole treatment leading to cell death.

We have treated 5 days old cells for 24h with 30 µg/ml Noc. We then washed the drug and transferred the cells into a glucose free medium. We then followed both cell survival, using methylene blue, and the cell’s capacity to form a colony after refeeding. In these conditions, we did not observe any toxic effect of the nocodazole. This result is now provided in Sup Fig 1L and discussed line 172-176.

(7) The "Tub1-only" mutant displays shorter but stable Q-nMT bundles in phase II, although they are thinner than in wild-type cells. What happens in the "Tub3-only" mutant, which also has beta-tubulin levels similar to wild-type cells (Supp. Fig. 2B)?

In order to measure Q-nMT bundle length and thickness, we used Tub1 fused to GFP. This cannot be done in a Tub3-only mutant. Yet, we have measured Q-nMT bundle length in Tub3-only cells using Bim1-3GFP as a MT marker (as in Laporte et al, JCB 2013). As shown in the figure below, Q-nMT bundles were shorter in Tub3-only cells than in WT cells whatever the phase.

We do not know if this effect is directly linked to the absence of Tub1 or if it is very indirect and for example due to the fact that Tub1 and Tub3 interact differently with Bim1 or other proteins that are involved in Q-nMT bundle stabilization. As we cannot give a clear interpretation for that result, we decided not to present those data in our manuscript.

(8) Why were wild-type and ndc80-1 cells imaged after a 20 min nocodazole treatment to evaluate the role of KT-MT attachments in Q-nMT bundle formation (Fig 3A)? Importantly, this experiment is also missing a control in which Q-nMT length is analyzed in both wild-type and ndc80-1 cells at 25ºC instead of 37ºC.

In this experiment, we used nocodazole to test both the formation and the stability of the Q-nMT bundle. Fig 3A shows MT length distribution in WT (grey) and ndc80-1 (violet) cells expressing mTQZTub1 (green) and Nuf2-GFP (red), shifted to 37 °C at the onset of glucose exhaustion and kept at this non-permissive temperature for 12 or 96 h then treated with Noc. The control experiment was provided in Sup Fig 3B. Indeed, this figure shows MT length in WT (grey) and ndc80-1 (violet) expressing mTQZ-Tub1 (green) and Nuf2-GFP (red) grown for 4 d (96h) at 25 °C, and treated or not with Noc. This is now indicated in the text line 216 and in the figure legend line 976

(9) As a general comment linked to the previous concern, it is striking that in many instances, Q-nMT bundle length is measured after nocodazole treatment without any evident reason to do this and without displaying the results in untreated cells as a control. If nocodazole is used, the authors should explicitly indicate it and state the reason for it.

We provide control experiments without nocodazole for all of the figures. For the sake of figure clarity, for Fig.3A the control without the drug is in Sup. Fig. 3B, for Fig. 3B it is shown in Sup. Fig. 3D, for Fig. 4B, it is shown in Sup. Fig 4A. This is now stated in the text and in the figure legend: for Fig. 3A: line 216 and in the figure legend line 976; for Fig. 3B: line 222 and figure legend line 984; for Fig. 4B: line 280 and in the figure legend line 1017.

The only figures where the untreated cells are not shown is for Fig 1D since the goal of the experiment is to make dynamic MTs shorten.

In Fig. 5C and Sup. Fig. 5D to F, we used nocodazole to get rid of dynamic cytoplasmic MTs that form upon quiescence exit in order to facilitate Q-nMT bundle measurement. This was explained in our previous study (Laporte et al, JCB 2013). We now mention it in the figure legends, see for example Fig. 5 legend line 1054.

(10) Ipl1 inactivation using the ipl1-1 thermosensitive allele impedes Q-nMT bundle formation. The inhibitor-sensitive ipl1-as1 allele could have been further used to show whether this depends on its kinase activity, also avoiding the need to increase the temperature, which affects MT dynamics. As suggested, we have used the ipl1-5as allele. We have thus modified Fig 3B and now show that is it indeed the Ipl1 kinase activity that is required for Q-nMT bundle formation initiation (line 222). In any case, it is surprising that deletion of SLI15 does not affect Q-nMT formation (in fact, MT length is even larger), despite the fact that Sli15, which localizes and activates Ipl1, is present at the Q-nMT (Fig 3C). Likewise, deletion of BIR1 has barely any effect on MT length after 4 days in quiescence (Fig 3D). Do the previous observations mean that Ipl1 role is CPC-independent? Does the lack of Sli15 or Bir1 aggravate the defect in Q-nMT formation of ipl1-1 cells at non-permissive or semi-permissive temperature?

Thanks to the Reviewer’s comments, we have re-checked our sli15Δ strain and found that it was accumulating suppressors very rapidly. To circumvent this problem, we utilized the previously described sli15-3 strain (Kim et al, JCB 1999). We found that sli15-3 was synthetic lethal with both ipl1-1, ipl1-2 (as described in Kim et al, JCB 1999) and with ipl1-as5, preventing us from addressing the CPC dependence of the Ipl1 effect asked by the Reviewer. However, using the sli15-3 strain, we now show that inactivation of Sli15 upon glucose exhaustion does prevent Q-nMT bundle formation (See new Sup Fig 3F and the text line 226-227).

(11) Lack of both Bir1 and Bim1 act in a synergistic way with regard to the defect in Q-nMT bundle formation. Although the absence of both Sli15 and Bim1 is proposed to lead to a similar defect, this is not sustained by the data provided, particularly in the absence of nocodazole treatment (Supp. Fig 3E).

Deletion of bir1 alone has only a subtle effect on Q-nMT bundle length in the absence of Noc, yet in bir1Δ cells, Q-nMT bundles are sensitive to Noc. Deletion of BIM1 (bim1Δ) aggravates this phenotype (Fig. 3D). As mentioned above, Q-nMT bundle formation is impaired in sli15-3 cells. In our hands, and as expected from (Zimnaik et al, Cur Biol 2012), this allele is synthetic lethal with bim1Δ.

On the other hand, the simultaneous lack of Bir1 and Bim1 drastically reduces the viability of cells in quiescence and this is proposed to be evidence supporting that KT-MT attachments are critical for QnMT bundle assembly (Supp Fig 3G). However, similarly to what was indicated previously for the 24 h nocodazole treatment, here again, the lack of viability could be originated by other reasons that are associated with the lack of Bir1 and Bim1 and not necessarily with problems in Q-nMT formation. In fact, the viability defect of cells lacking Bir1 and Bim1 is similar to that of cells only lacking Bir1 (Supp Fig 3G).

We have previously shown that many mutants impaired for Q-nMT bundle formation (dyn1Δ, nip100Δ etc) have a reduced viability in quiescence (Laporte et al, JCB 2013). In the current study, a very strong phenotype is observed for other mutants impaired for Q-nMT bundle formation such as bim1Δ bir1Δ cells, but also for slk19Δ bim1Δ.

Importantly, as shown in the new Sup Fig 1L, in WT cells treated with Noc upon entry into quiescence, a treatment that prevents Q-nMT formation, showed a reduced viability, while a Noc treatment that does not affect Q-nMT bundle formation, i.e. a treatment in late quiescence, has no effect on cell survival. This solid set of data point to a clear correlation between the ability of cells to assemble a Q-nMT bundle and their ability to survive in quiescence. Yet, of course, we cannot formally exclude that in all these mutants, the reduction of cell viability in quiescence is due to another reason.

(12) Both Mam1 and Spo13 are, to my knowledge, meiosis-specific proteins. It is therefore surprising that mutants in these proteins have an effect on MT bundle formation (Fig 3G-H, Supp. Fig. 3G). Are Mam1 and Spo13 also expressed during quiescence? Transcription of MAM1 or SPO13 does not seem to be induced by glucose depletion in previously published microarray experiments, but if Mam1 are Spo13 are expressed in quiescent cells, the authors should show this together with their results.

Indeed, it is interesting to notice that Mam1 and Spo13 are involved in both meiosis and Q-nMT bundle formation. As suggested by the Reviewer we have performed western blots in order to address the expression of those proteins in proliferation and quiescence (4d). We tagged Spo13 with either GFP, HA or Myc but none of the fusion proteins were functional. Yet, as shown in the new Sup Fig 3I, Mam1-GFP, Csm1-GFP and Lsr4-GFP were expressed both in proliferation and quiescence.

(13) In the laser ablation experiments that demonstrate that KT-MT attachments are not needed in order to maintain Q-nMT bundles once formed, anaphase spindles of proliferating cells were cut as a control (Supp. Fig 3I). However, late anaphase cells have already segregated the chromosomes, which lie next to the SPBs (this can be evidenced by looking at Dad2-GFP localization in Supp. Fig 3I), so that only interpolar MTs are severed in these experiments. The authors should have instead used metaphase cells as a control, since chromosomes are maintained at the spindle midzone and the length and width of the metaphase spindle is more similar to that of the Q-nMT bundle.

We have tried to “cut” short metaphase spindles, but as they are < 1 µm, after the laser pulse, it is difficult to verify that spindles are indeed cut and not solely “bleached”. Furthermore, after the cut, the remaining MT structure that is detectable is very short, and we are not confident in our length measurements. Yet, this type of experiment has been done in S. pombe (Khodjakov et al, Cur Biol 2004 and Zareiesfandabadi et al, Biophys. J. 2022). In these articles the authors have demonstrated that after a cut, metaphase spindles are unstable and rapidly shrink through the action of Kinesin14 and dynein. This is now mentioned in the text line 265.

(14) In the experiment that shows that cycloheximide prevents Q-nMT disassembly after quiescence exit, and therefore that this process requires de novo protein synthesis (Fig. 5A), cells are indicated to express only Spc42-RFP and Nuf2-GFP. However, Stu2-GFP images are also shown next to the graph and, according to the figure legend, it was indeed Stu2-GFP that was used to measure individual QnMT bundles in cells treated with cycloheximide. In the graph, additionally, time t=0 represents the onset of MT bundle depolymerization, but Q-nMT bundle disassembly does not take place after cycloheximide treatment. The authors should clarify these aspects of the experiment.

Following the Reviewer’s suggestion, to clarify these aspects we have split Fig. 5A into 2 panels.

Finally, some minor issues are:

(1) The text should be checked for proper spelling and grammar.

We have done our best.

(2) In some instances, there is no indication of how many cells were imaged and analyzed.

We now provide all these details either in the figure itself or in the figure legend.

(3) Besides the Q-nMT bundle, it is sometimes noticeable an additional strong cytoplasmic fluorescent signal in cells that express mTQZ-Tub1 and/or mRuby-Tub1 (e.g., Figs 1F, 1H and, particularly, Supp Fig 1H). What is the nature of these cytoplasmic MT structures?

We did mention this observation in the material and methods section (see line 526-528). This signal is a background fluorescence signal detected with our long pass GFP filter. It is not GFP as it is “yellowish” when we view it via the microscope oculars. This background signal can also be observed in quiescent WT cells that do not express any GFP. We do not know what molecule could be at the origin of that signal but it may be derivative of an adenylic metabolite that accumulates in quiescence and could be fluorescent in the 550nm –ish wavelength, but this is pure speculation.

(4) It is remarkable that a 20-30% decrease in tubulin levels had such a strong impact on the assembly of the Q-nMT bundle (Supp. Fig. 2). Can this phenotype be recovered by increasing the amount of tubulin in the mutants impaired for tubulin folding?

Yes, this is astonishing, but we believe our data are very solid since we observed that with both tub3Δ and in all the tubulin folding mutants we have tested (See Sup. Fig. 2). To answer Reviewer’s question, we would need to increase the amount of properly folded tubulin, in a tubulin folding mutant. One way to try to do that would be to find suppressors of GIM mutations, but this is a lengthy process that we feel would not add much strength to this conclusion.

(5) The graphs displaying the length of the Q-nMT bundle in several mutants in microtubule motors throughout a time course are presented in a different manner than in previous experiments, with data points for individual cells being only shown for the most extreme values (Fig 4C, 4H). It would be advisable, for the sake of comparison, to unify the way to represent the data.

We have now unified the way we present our figures.

(6) How was the exit from quiescence established in the experiments evaluating Q-nMT disassembly? How synchronous is quiescence exit in the whole population of cells once they are transferred to a rich medium?

We set the “zero” time upon cell refeeding with new medium. In fact, quiescence exit is NOT synchronous. We have reported this in previous publications, with the best description of this phenomena being in Laporte et al, MIC 2017 .

The figures below are the same data but on the left graph, the kinetic is aligned upon SPB separation onset, while on the right graph (Fig 5A), it is aligned on MT shrinking onset.

We can add this piece of data in a Sup Figure if the Reviewer believes it is important.

Reviewer #2 (Recommendations For The Authors):

General:

In general, more precise language that accurately describes the experiments would improve the text.

We have tried to do our best to improve the text.

The authors should clearly define what they mean by an active process and provide context to support this statement regarding the Q-nMT.

We have strived to clarify this point in the text (see paragraph form line 146 to 178).

It is reasonable to assume that structures composed of microtubules are dynamic during the assembly process. The authors should clarify what they mean by "stable by default i.e., intrinsically stable." Do they mean that when Q-nMT assembly starts, it will proceed to completion regardless of a change in condition?

We mean that in phase I the Q-nMT bundle is stabilized as it grows and that stabilization is concomitant with polymerization. By contrast, MTs polymerized during phase II are not stabilized upon elongation beyond the phase I polymer, and get stabilized later, in a separate phase (i.e. in phase III). We hope to have clarified this point in the text (see line 108-110).

In lines 33-34, the authors claim that the Q-nMT bundle functions as a "sort of checkpoint for cell cycle resumption." This wording is imprecise, and more significantly the authors do not provide evidence supporting a direct role for Q-nMT in a quiescence checkpoint that inhibits re-entry into the cell cycle.

We have softened and clarified the text in the abstract (see line 29-30)., in the introduction (line 101104), in the result section (line 331-332) and in the discussion (line 426-430).

Many statements are qualitative and subjective. Quantitative statements supported by the results should be used where possible, and if not possible restated or removed.

We provide statistical data analysis for all the figures.

The number of hours after glucose exhaustion used for each phase varies between assays. This is likely a logistical issue but should be explained.

This is indeed a logistical issue and when pertinent, it is explained in the text.

It would be interesting to address how this process occurs in diploids. Do they form a Q-nMT? How does this relate to the decision to enter meiosis?

Diploid cells enter meiosis when they are starved for nitrogen. Upon glucose exhaustion diploids do form a Q-nMT bundle. This is shown and measured in the new Sup Fig1C. In fact, in diploids, Q-nMT bundles are thicker than in haploid cells.

It would be interesting to address how the timescale of this process compares to the types of nutrient stress yeast would be exposed to in the environment.

We have transferred proliferating yeast cells to water, to try to mimic what could happen when yeast cells face rain in the wild. As shown below, they do form a Q-nMT bundle that becomes nocodazole resistant after 30h. This data is now provided in the new Sup Fig 1D.

It is recommended that the authors use FRAP experiments to directly measure the stability of the QnMT bundles.

This experiment was published in (Laporte et al, 2013). Please see response to Reviewer #1.

In many cases, the description of the experimental methods lacks sufficient detail to evaluate the approach or for independent verification of results.

We have strived to provide a more detailed material and methods section, as well as more detailed figure legends and statistical informations.

Specific comments on figures:

In Figure 1 c), what do the polygons represent? They do not contain all the points of the associated colour.

The polygon represented the area of distribution of 90% of the data points. As they did not significantly add to the data presentation they have been removed.

In Figure 2 a), is the use of two different sets of markers to control for the effect of the markers on microtubule dynamics?

Yes, we are always concerned about the influence of GFP on our results, so very often we replicate our experiments with different fluorescent proteins or even with different proteins tagged with GFP. This is now mentioned in the text (line 184-186).

Is it accurate to say (line 201, figure 3 a)) that no Q-nMT bundles were detected in ndc80-1 cells shifted to 37 degrees, or are they just shorter?

As shown in Fig 3A, in ndc80-1 cells, most of the MT structures that we measured are below 0,5um. This has been re-phrased in the text (line 214-215).

Lines 265-269, figure 4 b), how can the phenotype observed in cin8∆ cells be explained given the low abundance of Cin8 that is detected in quiescent cells?

Faint fluorescence signal is not synonymous of an absence of function. As shown in Sup Fig 4B, we do detect Cin8-GFP in quiescent cells.

Quantification is needed in Figure 4 panels c) and h).

Fig 4C and 4H have been changed and quantification are provided in the figure legend.

Reviewer #3 (Recommendations For The Authors):

A few points should be addressed for clarity:

(1) Sup. Fig. 1K: are only viable cells used for the colony-forming assay? How were these selected? If not, the assay would just measure survival (as in the viability assay).

Yes, only viable cells were selected for the colony forming assay. We used methylene blue to stain dead cells. Then, we used a micromanipulation instrument (Singer Spore Play) that is commonly used for tetrad dissection to select “non blue cells” and position them on a plate (as we do with spores). Each micromanipulated cell is then allowed to grow on the plate and we count colonies (see picture in Sup Fig 1L right panel). This was described in Laporte et al, JCB 2011. We have added that piece of information in the legend (line 1129-1130) and in the M&M section (line 580-586).

(2) Could Tub3 have a role in phase I? It is not clear why the authors conclude involvement only in phase II.

As it can be seen in Fig 2D, MT bundle length and thickness are quite similar in WT and Tub1-only cells in phase I, indicating that the absence of Tub3 as no effect in phase I. In Tub1-only cells, MT bundles are thinner in both phase II and phase III, yet, they get fully stabilized in phase III. Thus, the effect of Tub3 is largely specific to the nucleation/elongation of phase II MTs. We hope to have clarified that point in the text (line 203-207).

(3) Quantifications, statistics: for all quantifications, the authors should clearly state the number of experiments (replicates), and number of cells used in each, and what number was used for statistics. For all quantifications in cells, it seems that the values from the total number of cells across different experiments were plotted and used for statistics. This is not very useful and results in extremely small p values. I assume that the values for individual cells were obtained from multiple, independent experiments. Unless there are technical limitations that allow only a very small sample size (not the case here for most experiments), for experiments involving treatments the authors should determine values for each experiment and show statistics for comparison between experiments rather than individual cells pooled from multiple experiments.

All the experiments have been done at least in replicate. In the new Fig. 1A, we now display each independent experiment with a specific color code. For Fig 2B and 2C we now provide the data obtained for each separate experiment in Sup Fig 2C. Additional details about quantifications and statistics are provided in the M&M section or in the specific figure legends.

Version 2

Abstract

Abstract

Cells fine-tune microtubule assembly in both space and time, to give rise to distinct edifices with specific cellular functions. In proliferating cells, microtubules are highly dynamics, and proliferation cessation often leads to their stabilization. One of the most stable microtubule structures identified to date is the nuclear bundle assembled in quiescent yeast. In this report, we characterize the original multistep process driving the assembly of this structure. This AuroraB-dependent mechanism follows a precise temporality that relies on the sequential actions of kinesin-14, kinesins-5 and involves both microtubule-kinetochore and kinetochore-kinetochore interactions. Upon quiescence exit, the microtubule bundle is disassembled via a cooperative process involving kinesin-8 and its full disassembly is required prior to cells re-entry into proliferation. Overall, our study provides the first description, at the molecular scale, of the entire life cycle of a stable microtubule structure in vivo, and sheds light on its physiological function.

Introduction

Microtubules (MTs) are hollow cylindrical polymers assembled by the non-covalent interaction of α- and β-tubulin heterodimers. MTs are generally nucleated at MT organizing centers (MTOC) by a ɣ-tubulin complex (ɣ-TuC) which acts as a template and stabilizes the so-called MT “minus-end” (Liu et al., 2021; Roostalu and Surrey, 2017; Sanchez and Feldman, 2017; Thawani and Petry, 2021). At the opposite end, the “plus-end” (+end), MTs elongate by the addition of GTP tubulin. The β-tubulin bound GTP is hydrolyzed, and stable but transient, GDP + Pi intermediates are generated in the region immediately behind the growing +end. Subsequent Pi release favors MT depolymerization which can be rescued by de novo GTP-tubulin addition (Cleary and Hancock, 2021; Gudimchuk and McIntosh, 2021). Thus, MTs alternate between periods of growth and shrinkage, a behavior termed dynamic instability (Mitchison and Kirschner, 1984). In vivo, a profusion of MT-associated proteins (MAPs) regulate MT length and dynamics (Bodakuntla et al., 2019; Goodson and Jonasson, 2018). In addition, MTs are often assembled from multiple tubulin variants or isotypes, and are modified by a cohort of post-translational modifications that modulate MT dynamics directly or by influencing the recruitment of MAPs (Janke and Magiera, 2020; Roll-Mecak, 2020). Other MAPs organize MTs into highly-ordered assemblies either by cross-linking MTs or by connecting them with cellular structures such as membranes, chromosomes, or actin cytoskeleton components (Bodakuntla et al., 2019; Meiring et al., 2020). Cells fine-tune these mechanisms in both space and time to give rise to distinct MT edifices with specific functions.

In proliferating cells, MT dynamics are crucial for their functions. It allows exploration of the cell volume in search of structures to “capture”, such as centromeres during mitosis, or to exert pushing or pulling forces required for various cellular processes such as cell migration or morphogenesis (Heald and Khodjakov, 2015; Kirschner and Mitchison, 1986). The MT cytoskeleton can undergo extensive rearrangements and assemble more stable MT structures, especially when cells change fate (Meiring et al., 2020; Röper, 2020). For example, in terminally differentiated cells, such as epithelial cells, cardiomyocytes or neurons, stable MT networks make up the majority, and ensure critical cellular functions such as cell shape maintenance or long-distance intracellular transport (Baas et al., 2016; Muroyama and Lechler, 2017). Defects in MT stabilization are at the origin of many human pathologies, including neurodegenerative diseases and ciliopathies (Anvarian et al., 2019; Wheway et al., 2018). Effectors responsible for MT stabilization have been actively searched for since the 1980’s. In mammals, the involvement of specific MAPs such as Tau, MAP2, MAP6 and PRC1, or tubulin post-translational modifications has been extensively studied. While their contributions to MT stabilization are central and undisputed, their sole actions do not fully explain the observed levels of MT stability in many types of non-dividing cells (Hahn et al., 2019).

For years, yeast species have been instrumental in unraveling mechanisms that regulate MT dynamics in eukaryotes. While proliferating yeast cells exhibit a dynamic MT network (Winey and Bloom, 2012), proliferation cessation goes with the formation of dramatically stable MT structures in both S. cerevisiae and S. pombe (Laporte and Sagot, 2014; Laporte et al., 2013, 2015). Upon quiescence establishment, S. cerevisiae cells assemble a bundle composed of stable parallel MTs, hereafter referred to as the Q-nMT bundle, for Quiescent-cell nuclear Microtubule bundle. This structure originates from the nuclear side of the spindle pole body (SPB), the yeast equivalent of the centrosome. It spans the entire nucleus and relocalizes kinetochores and centromeres, which remain attached to MT +ends (Laporte and Sagot, 2014; Laporte et al., 2013). Because the MTs embedded in the Q-nMT bundle are not all of the same length, the structure has a typical arrow shape. When cells exit quiescence, the Q-nMT bundle depolymerizes and, by pulling the attached centromeres back to the SPB, allows the recovery of the typical Rabl-like configuration of chromosomes found in G1 yeast cells (Jin et al., 1998). The molecular mechanisms underlying the formation of this peculiar stable MT structure and its physiological function(s) are not understood. However, cells defective in Q-nMT bundle formation have a compromised survival in quiescence and a drastically reduced fitness upon cell cycle re-entry (Laporte and Sagot, 2014; Laporte et al., 2013, 2015).

Here, we show that Q-nMT bundle formation is a multistep process that follows a precise temporal sequence. The first step relies on AuroraB/Ipl1 activity and requires the kinesin-14 Kar3, its regulator Cik1, and the EB1 homolog Bim1. It leads to the formation of a short (≈ 1 µm) and stable bundle that resembles a half mitotic spindle. Importantly, in this first step, MT polymerization and stabilization are coupled. In a second step, additional MTs polymerize from the SPB, elongate, and are zipped to pre-existing MTs in a Cin8/kinesin-5 dependent manner. Complete stabilization of the Q-nMT bundle is achieved in a third and final step that requires the kinesin-5 Kip1. Our observations further indicate that Q-nMT bundle assembly requires both MT-kinetochore attachment and inter-kinetochore interactions. Finally, we show that, upon exit from quiescence, the Q-nMT bundle disassembles from its +ends, each MT depolymerizing in coordination with its neighbors, via the action of the depolymerase Kip3, a member of the kinesin-8 family. Importantly we show that the complete disassembly of the Q-nMT bundle is required for cells to re-enter the proliferation cycle. Overall, this study describes the entire life cycle of a MT structure, from the molecular mechanisms involved in its formation and stabilization to its disassembly and further suggests that this atypical quiescent specific structure acts as a regulator, or “control point”, for cell cycle resumption upon exit from quiescence.

Results

The Q-nMT bundle formation is a three-step process

We have previously shown that the Q-nMT bundle observed in quiescent cells is a stable nuclear structure (Laporte et al., 2013). To investigate whether the stabilization of the Q-nMT bundle was concomitant with its assembly or whether it assembled first as a dynamic structure, and then gets stabilized, we quantified the Q-nMT bundle length (Fig. 1A) and thickness (Fig. 1B) upon quiescence establishment following carbon source exhaustion. We found that the Q-nMT bundle assembly process could be divided into 3 sequential phases. In an initial phase (phase I), MTs elongated from the SPB to reach ≈0.8 µm. The number of these MTs, referred to as phase I-MTs, was approximately the same as in a mitotic spindle (see inset of Fig. 1B and Sup. Fig. 1A). Importantly, phase I-MTs were resistant to nocodazole (Noc), a MT poison that causes dynamic MT depolymerization, indicating that phase I-MT stabilization was concomitant with their polymerization (Fig. 1A-B). In a second phase, beginning ≈10h after glucose exhaustion, additional MTs emerged from the SPB (phase II-MTs) and elongated along phase-I MTs, nearly doubling the thickness of the phase I MT bundle (Fig. 1B and Sup. Fig. 1A). At this stage, the tip of the newly elongated MTs were unstable (Fig. 1A-B). Complete stabilization of the Q-nMT bundle was achieved ≈48h after glucose exhaustion (phase III). Indeed, after this step, a Noc treatment did not affect either the length or the thickness of the Q-nMT bundle (Fig. 1A-B and Sup. Fig. 1B). Plotting Q-nMT bundle width as a function of length for each individual cell, before and after Noc treatment further revealed the 3 successive phases of the Q-nMT bundle formation (Fig. 1C). Of note, Q-nMT bundles were also observed both in diploid cells upon glucose exhaustion (Sup. Fig. 1C) and in haploid cells transferred to water (Sup. Fig. 1D).

The formation of the Q-nMT bundle is a three-step process

(A) Nuclear MT length in WT cells expressing mTQZ-Tub1, before (-) or after (+) a 15 min Noc treatment (30 µg/ml) upon entry into quiescence. Each circle corresponds to the length of an individual MT structure. The results for 3 independent experiments are shown (pale blue, cyan and dark blue, n > 160 for each point in each experiment). The mean and SD are shown. Student test or ANOVA (sample >2) were used to compare inter-replicates. #: p-value > 0.05, $: p-value < 0.05. A student test (t-test with two independent samples) was used to compare the results obtained with or without Noc, the indicated p-values being the highest measured among experiments. ***: p-value < 1.10−5. Images of representative cells are shown. Bar is 2 µm.

(B) MT fluorescence intensity as a proxy of MT structure width in WT cells expressing mTQZ-Tub1. Mean intensity measurement for half pre-anaphase mitotic spindles (purple), phase I (green), phase II (orange) or phase III (red) Q-nMT bundle. A line scan along the MT structure for an individual cell is shown as thin line, the mean as a bold line (n > 60 / phase), all the lines being aligned at 0,5 µm before the fluorescence intensity increase onset on the SPB side. The blue lines are results obtained after a 15 min Noc treatment (30 µg/ml). In each graph, horizontal dashed line indicates the mean intensity. Images in pseudo-colors of a representative cell for each phase are shown. Bar is 2 µm.

(C) MT bundle length as a function of MT bundle width for individual cells in each phase before and after a 15 min Noc treatment (30 µg/ml) in WT cells expressing mTQZ-Tub1. Each circle represents an individual MT structure.

(D) WT cells expressing mTQZ-Tub1 (red) and Nuf2-GFP (green) in phase II (23 h) or phase III (50 h) were deposited on an agarose pad containing 30 µg/ml Noc and imaged. Blue arrowheads: SPB; white arrowheads: Nuf2-GFP clusters. Time is in min after deposition on the pad. Bar is 1 µm.

(E) Tub4-mTQZ fluorescence intensity measured at the SPB upon entry into quiescence. Each circle represents an individual cell. The mean and SD are shown; t-tests were used to compare independent samples (N = 3, n > 150), *: 0.05 > p-value > 1.10−3, ***: p-value < 1.10−5. Images in pseudo-colors of representative cells are shown. Bar is 2 µm.

(F) WT cells expressing mTQZ-Tub1 under the TUB1 promoter and mRuby-Tub1 under the ADH2 promoter. The percentage of cells harboring both mTQZ and mRuby fluorescence along the Q-nMT bundle is shown; each circle being the percentage for an independent experiment, with n > 200 counted for each experiment. The mean and SD are shown. Images of representative cells at the indicated time after glucose exhaustion are shown. Bar is 2 µm.

(G) Schematic of the Q-nMT bundle formation. During phase I, stable MTs (phase I MT - green) elongate from the SPB (grey). During phase II, the amount of Tub4 (cyan) increases at the SPB. In the meantime, new MTs (phase II MT - orange) elongated from the SPB and are stabilized along the phase I MTs, yet their +ends remain dynamic (dashed lines). After phase III, all MTs are stabilized (red). Nuf2 is schematized as dark blue squares.

(H) Upon glucose exhaustion, WT cells expressing mTQZ-Tub1 (green) and Nuf2-GFP (red) were pulsed treated with 30 µg/ml Noc (blue) or DMSO (grey) for 24 h. Noc or DMSO were then chased using carbon exhausted medium and cells were imaged. Each circle corresponds to MT structure length in an individual cell. The mean and SD are shown (N = 3, n > 100). Images of representative cells 2 d after the chase and representative cells 4 days after the chase are shown. Tubulin (green) was detected by immunofluorescence, actin (red) by phalloidin and DNA (blue) with DAPI. The mean Q-nMT bundle length (±SD) in the population is indicated.

To confirm the above observations, first, we followed Nuf2-GFP, a protein that localizes to the MT +end. As expected, movies showed that upon Q-nMT bundle formation initiation, Nuf2-GFP signal relocated from the SPB to the distal part of elongating Q-nMT bundles (Sup. Fig. 1E). In phase II, when cells were treated with Noc, the Nuf2-GFP signal followed depolymerizing MTs (Fig. 1D, left panel), testifying for the instability of phase II-MTs. In contrast, the Nuf2-GFP signal remained immobile upon Noc treatment in phase III, as cells have fully stabilized Q-nMT bundles (Fig. 1D, right panel). Second, we measured the γ-tubulin (Tub4) signal at the SPB and found that Tub4 started to accumulate at the SPB at the beginning of phase I to reach a plateau at the end of phase II (Fig. 1E and Sup. Fig. 1F). Interestingly, the amount of Tub4 at the SPB was proportional to the thickness of the Q-nMT bundle (Sup. Fig. 1G). Finally, in order to confirm the two waves of MT elongation, we developed a strain expressing mTQZ-Tub1 from the endogenous TUB1 promoter and mRuby-TUB1 under the ADH2 promoter, a promoter that is derepressed after glucose exhaustion, as confirmed by western blot (Sup. Fig. 1H). We observed that mRuby-TUB1 was incorporated into the Q-nMT bundle only after phase I (Fig. 1F and Sup. Fig. 1H), confirming the existence of a second wave of MT elongation during phase II. All the above experiments led us to propose the model shown in Figure 1G.

Q-nMT bundle formation is an essential time-regulated process

Increasing cytoplasmic viscosity has been shown to dampen MT dynamics (Molines et al., 2022). Upon quiescence establishment, yeast cells undergo a transition from a fluid phase to a solid-like phase (Munder et al., 2016; Joyner et al., 2016) and an acidification (Jacquel et al., 2021) which could contribute to the formation of the Q-nMT bundle. Several lines of evidence suggest that these changes in physicochemical properties are not involved in Q-nMT formation. First, 4-day-old quiescent cells can simultaneously display both dynamic cytoplasmic MTs (cMTs) and a stable Q-nMT bundle (Sup. Fig. 1I and (Laporte et al., 2013)). Second, when we mimicked a fluid phase to solid-like transition in proliferating cells by artificially lowering the pH, no stable MTs were observed, whereas F-actin aggregation was induced (Sup. Fig. 1J) as expected (Peters et al., 2013). Third, 2-day-old quiescent cells exhibiting a phase III Q-nMT bundle were able to grow de novo dynamic cMTs after a pulse-chase Noc treatment (Sup. Fig. 1K).

Importantly, if Q-nMT bundle formation was solely dependent on physicochemical changes, this structure should be able to assemble in late quiescence. By treating cells with Noc upon glucose exhaustion, we were able to prevent the formation of the Q-nMT bundle in early quiescence (Fig. 1H). Strikingly, when the drug was washed out, no Q-nMT bundle assembled, even 144h after Noc removal (Fig. 1H). These cells had entered quiescence because they had assembled actin bodies (Fig. 1H, right panel), another quiescent cell-specific structure (Sagot et al., 2006). This experiment strongly suggests that Q-nMT formation is a process induced by a transient signal emitted upon glucose exhaustion.

We have previously shown that mutants unable to assemble Q-nMT bundles have reduced viability in quiescence and a reduced ability to form colonies upon exit from quiescence (Laporte et al., 2013). To establish a direct link between the absence of the Q-nMT bundle and the above phenotypes, we took advantage of our ability to conditionally prevent Q-nMT bundle formation in WT cells using Noc treatment at the onset of glucose exhaustion. As shown in Sup. Fig. 1L, in the absence of the Q-nMT bundle, WT cells lose viability in quiescence and survivors have a reduced ability to generate a progeny upon exit from quiescence. In contrast, a similar Noc treatment of cells that have already assembled a stable Q-nMT bundle (5-day-old cells) did not affect either the viability in quiescence or quiescence exit efficiency, demonstrating that Noc does not have a toxic effect per se. Taken together, these experiments confirm our initial findings that the Q-nMT bundle is important for both cell survival during chronological aging and quiescence exit fitness.

Steady state tubulin levels and Tub3 isotype are critical for Q-nMT bundle formation

We then focused on deciphering the molecular mechanisms involved in Q-nMT bundle formation. Yeast cells express two α-tubulin isotypes, Tub1 and Tub3. In proliferating cells, Tub1 constitutes the majority (Aiken et al., 2019; Nsamba et al., 2021; Schatz et al., 1986). In 4-day-old cells, the two α-tubulin isotypes were found embedded in the Q-nMT bundle. This was observed using two different pairs of fluorescent proteins in order to avoid any influence of the FP variants, using both widefield and by expansion microscopy (Fig. 2A and Sup. Fig. 2A). Since Tub3 stabilizes MTs in vitro (Bode et al., 2003), it may be key for Q-nMT bundle formation. To test this hypothesis, we first analyzed the phenotype of tub3Δ cells. In phase II, tub3Δ cells displayed short MT bundles that were sensitive to Noc. In phase III, no MT bundles were detected (Fig. 2B and Sup. Fig. 2C). Thus, in this mutant, the MT bundles assembled during phase I were not stable and eventually collapsed. In tub3Δ cells, the steady-state amount of α-tubulin is significantly reduced (Sup. Fig. 2B and (Nsamba et al., 2021)). Thus, the amount of α-tubulin, but not Tub3 itself, may be important for Q-nMT bundle formation. Indeed, mutants defective in α- and β-tubulin folding, such as the prefolding complex mutants pac10Δ, gim3Δ or yke2Δ, or in tubulin heterodimer formation, such as pac2Δ, or in β-tubulin folding, such as cin2Δ or cin4Δ, all of which have a reduced amount of tubulin, were unable to assemble a Q-nMT bundle (Sup. Fig. 2D-E).

Q-nMT bundle formation is influenced by the alpha-tubulin amount and isotype

(A) WT cells (4 d) expressing either Tub3-3GFP (green) and mTQZ-Tub1 (red, top panel) or Tub3-RFP (red) and mWasabi-Tub1 (green, bottom panel). Blue arrowheads point to SPB, bar is 2 µm.

(B) Nuclear MT length in WT and tub3Δ cells expressing mTQZ-Tub1, 36 h (phase II, orange) and 90 h (phase III, red) after glucose exhaustion, treated 15 min (blue) or not (grey) with 30 µg/ml Noc. Each circle corresponds to the length of an individual MT structure, mean and SD are shown. ANOVA was used to compare inter-replicates (n > 200, N = 5); #: p-value > 0.05, $: p-value < 0.05. A student test (t-test with two independent samples) was used to compare (+) or (-) Noc data. The indicated p-values are the highest calculated among the 5 experiments; ***: p-value < 1.10−5. Images of representative cells are shown, bar is 2 µm.

(C) Nuclear MT length in WT and Tub1-only cells expressing mTQZ-Tub1, 36 h (phase II, orange) and 90 h (phase III, red) after glucose exhaustion, treated 15 min or not with 30 µg/ml Noc. Statistical representations are as in (B). Images of representative cells are shown, bar is 2 µm.

(D) Fluorescence intensity along the Q-nMT bundle in WT (grey) and Tub1-only (blue) cells expressing mTQZ-Tub1 grown for 10 h (phase I), 36 h (phase II), and 90 h (phase III). Individual fluorescent intensity is shown as thin line, the mean as a bold line (n > 60 / phase), all the lines being aligned at 0,5 µm before the fluorescence intensity increase onset on the SPB side. Dashed lines indicate the maximal mean fluorescence intensity. Images in pseudo-color of representative cells are shown, bar is 2 µm.

To identify the role of Tub3 per se, we took advantage of the “Tub1-only” mutant, in which the TUB3 gene was replaced by the TUB1 gene at the TUB3 locus. This mutant expresses only Tub1 and has an α-tubulin level similar to WT (Sup. Fig. 2B and (Nsamba et al., 2021)). We found that “Tub1-only” cells have unstable Q-nMT bundles in phase II, and shorter yet stable Q-nMT bundles in phase III (Fig. 2C and Sup. Fig2. C). Moreover, “Tub1-only” Q-nMT bundles were thinner than WT Q-nMT bundles (Fig. 2D). Taken together, our data indicate that Tub3 is involved in MT elongation in phase II, since in the absence of this isotype, phase I MT bundles could assemble and MT bundles could get stabilized in phase III, but the second wave of MT nucleation and elongation was impaired as testified by thinner Q-nMT bundles in both phase II and III.

Kinetochores are critical for phase I but dispensable for the maintenance of Q-nMT bundles

Kinetochore components are enriched at the Q-nMT bundle +end ((Laporte et al., 2013) and Sup. Fig. 3A). We therefore tested whether kinetochore-MT attachment might play a role in Q-nMT bundle formation. First, we disrupted kinetochore-MT attachment upon glucose exhaustion using the well-established ndc80-1 allele (Cheeseman et al., 2006; DeLuca et al., 2018; Wigge et al., 1998). As shown in Fig. 3A, only very short MT structures were detected in ndc80-1 cells transferred to 37°C at the onset of quiescence entry, even after 4 days at non-permissive temperature (for control see Sup. Fig. 3B). Second, we focused on the chromosome passenger complex (CPC: Bir1/Survivin, Sli15/INCENP, Nbl1/Borealin and Ipl1/Aurora B), a complex known to regulate kinetochore-MT attachment dynamics (Cairo and Lacefield, 2020). We found that inactivation of IPL1 upon entry into quiescence using the thermosensitive allele ipl1-1 prevented Q-nMT bundle formation (Sup. Fig. 3C). The absence of Q-nMT bundles in cells harboring the NA-PP1 kinase-sensitive allele ipl1-5as indicated that the Ipl1 kinase activity was required for this process (Fig. 3B and Sup. Fig. 3D-E for controls). While we could not detect Ipl1 and Nbl1 in quiescent cells, we found that Sli15-GFP and Bir1-GFP localized along the Q-nMT bundle with an enrichment at the Q-nMT bundle +end (Fig. 3C). In 4-day-old bir1Δ cells, Q-nMT bundle were shorter and not fully stabilized (Fig. 3D). In addition, inactivation of SLI15 upon glucose exhaustion using the thermosensitive allele sli15-3 drastically impaired Q-nMT bundle formation (Sup. Fig. 3F). Finally, we examined the involvement of Bim1. Bim1 is a MT +end binding protein that plays a role in kinetochore MT-end on attachment (Dudziak et al., 2021; Thomas et al., 2016) and localizes along the entire Q-nMT bundle (Laporte et al., 2013). In fact, we found that the amount of Bim1 correlated with tubulin incorporation during Q-nMT bundle formation (Sup. Fig. 3G). Bim1 deletion alone has no effect on Q-nMT bundle formation even in the presence of Noc ((Laporte et al., 2013) and Fig. 3D). However, bim1Δ bir1Δ cells barely assembled MT structures, and these structures were sensitive to Noc (Fig. 3D). This demonstrates that in the absence of Bim1, Bir1 is critical for Q-nMT bundle stabilization in phase I. Of note, we found that the viability of bim1Δ bir1Δ cells in quiescence was drastically reduced (Sup. Fig. 3H). Together, these experiments show that kinetochore-MT attachments are critical for the initiation of Q-nMT bundle formation upon entry into quiescence.

Kinetochore-kinetochore interactions are required for Q-nMT bundle formation

(A) Nuclear MT length distribution in WT (grey) and ndc80-1 (violet) cells expressing mTQZ-Tub1 (green) and Nuf2-GFP (red), transferred to 37 °C upon glucose exhaustion and maintained at 37 °C for the indicated time. Cells were imaged after a 20 min Noc treatment (30 µg/ml). Each circle corresponds to the length of an individual MT structure, mean and SD are shown. A student test (t-test with two independent samples) was used to compare (+) or (-) Noc samples from the same experiment (n > 250, N = 2). The indicated p-values are the highest p-values calculated among the 2 repeated experiments. ***: p-value < 1.10−5. Images of representative cells incubated 12 h at 37 °C are shown, bar is 2 µm.

(B) WT (grey) and ipl1-5as (pink) cells expressing mTQZ-Tub1 were treated for 60 h with 50 µM NA-PP1 after glucose exhaustion, and imaged after a 15 min Noc treatment (30 µg/ml). Same statistical representation as in (A); N = 2, n > 250. Images of representative cells are shown, bar is 2 µm.

(C) WT cells (2 d) expressing mTQZ-Tub1 (red) and Bir1-GFP or Sli15-GFP (green) were imaged. Graphs show Bir1-GFP or Sli15-GFP fluorescence intensity along normalized Q-nMT bundles (see material and method section, plain and dash lines: mean and SD respectively). Images of representative cells are shown, bar is 2 µm.

(D) Nuclear MT length distribution in cells of the indicated genotype (4 d) expressing mTQZ-Tub1 treated or not with Noc (30 µg/ml). Same statistical representation as in (A); N = 3, n > 200. Images of representative cells are shown, bar is 2 µm.

(E) WT cells expressing mTQZ-Tub1 (red) and Slk19-GFP (green) 4 h after glucose exhaustion. Blue arrowhead: SPB, bar is 2 µm.

(F) Nuclear MT length distribution in cells of the indicated genotype (4 d) expressing mTQZ-Tub1 (green) and Nuf2-GFP (red) imaged before or after Noc treatment (30 µg/ml). Same statistical representation as in (A); N = 2, n > 200. Images of representative cells are shown, white arrowheads point to Nuf2-GFP dots, bar is 2 µm.

(G) Cells of the indicated genotype (4 d) expressing mTQZ-Tub1 (green) and Nup2-RFP (red); bar is 1 µm.

(H) Nuclear MT length distribution in cells of the indicated genotype (4 d) expressing mTQZ-Tub1 treated or not with Noc (30 µg/ml). Same statistical representation as in (A); N ≥ 2, n > 200. Images of representative cells are shown, bar is 2 µm.

(I) Length variation of nuclear MT bundle fragments after laser ablation (pink dash line) in cells expressing mRuby-TUB1 (red) and Dad2-GFP (green). Time is in min. Images of representative cells are shown, blue arrowhead: SPB, white arrowhead: cMT, bar is 2 µm. Graph indicates the variation of length in released fragments (n > 120) and histograms show the % of released or SBP attached fragments that are either stable, or that shorten or grow within a 30 min period after the laser-induced breakage. Bar is 1 µm

We then hypothesized that inter-kinetochore interactions could stabilize parallel MTs embedded in the Q-nMT bundle. To test this idea, we analyzed MT organization in slk19Δ cells which are defective in kinetochore clustering (Richmond et al., 2013). In quiescence, Slk19-GFP was enriched at both ends of the Q-nMT bundles (Fig. 3E). slk19Δ cells displayed shorter and thinner MT structures that were sensitive to Noc (Fig. 3F). Since Bim1 has been implicated in kinetochore-kinetochore interaction, we combined slk19Δ with bim1Δ. MT structures detected in slk19Δ bim1Δ cells barely reached ≈1 µm and disappeared upon Noc treatment. In addition, in slk19Δ bim1Δ cells, kinetochores localized as a rosette around the SPB (Fig. 3F) and slk19Δ bim1Δ cell viability was severely impaired (Sup. Fig. 3H). This demonstrates that in the absence of Bim1, Slk19 is required for MT bundling and stabilization in phase I. We then focused on monopolin (Mam1, Lrs4, Hrr25 and Csm1), a complex that is known to cross-link kinetochores (Corbett et al., 2010)). These proteins were found to be expressed in quiescent cells (Sup Fig 3I) and when we analyzed the phenotype of mam1Δ, lrs4Δ and csm1Δ cells in quiescence, most of them displayed short MT structures often arranged in a star-like array (Fig. 3G-H and Sup. Fig. 3J). As for bir1Δ and slk19Δ, the combination of bim1Δ with monopolin deletion worsened MT structure length and stability (Fig. 3H) as well as cell viability in quiescence (Sup. Fig. 3H). Furthermore, cells deleted for SPO13, a gene encoding a protein involved in monopolin recruitment to kinetochores (Lee et al., 2004) showed similar phenotypes (Fig. 3G-H and Sup. Fig. 3H). Collectively, these experiments demonstrate that inter-kinetochore interactions are critical for Q-nMT bundle assembly.

Finally, we investigated whether kinetochore-MT interactions were required for the stability of the Q-nMT bundle once it is formed. In 5-day-old WT cells, we used a UV-pulsed laser to break the Q-nMT bundle into two pieces. In most cases, the length of both the released fragment and the fragment attached to the SPB remained constant, with the presence of dynamic cMTs after laser ablation testifying for cell viability (Fig. 3I, Sup. Fig. 3K). In contrast, as expected form previous studies (Khodjakov et al., 2004; Zareiesfandabadi and Elting, 2022), in proliferating cells, dynamic anaphase spindles promptly disassembled after the laser-induced breakage (Sup. Fig. 3L). This indicates that once Q-nMT bundles are formed, they do not require MT-kinetochore interaction to be maintained. Accordingly, inactivation of Ndc80 (Sup Fig. 3B) or Ipl1 (Sup. Fig. 3M) in cells that are already in quiescence had no effect on the Q-nMT bundle maintenance. Thus, once formed, Q-nMT stability is established and maintained throughout its length.

Each phase of Q-nMT bundle formation requires specific kinesins

To further dissect the molecular mechanism of Q-nMT bundle formation, we focused on kinesins. Kar3 is a kinesin that, in complex with its regulator Cik1, can generate parallel MT bundles from a MT organizing center (MTOC) both in vitro and in proliferating cells (Manning et al., 1999; Mieck et al., 2015; Molodtsov et al., 2016). In quiescence, we found that Kar3-3GFP localized to the SPB, but also as dots along the Q-nMT bundle (Fig. 4A). In 4-day-old kar3Δ cells, the majority of the detected MT structures were extremely short compared to WT (Fig. 4B, for control without Noc see Sup Fig. 4A). Similar results were obtained in cik1Δ but not in vik1Δ cells, which lack the alternative Kar3 regulators (Fig. 4B). Thus, the Kar3/Cik1 complex is required for phase I.

Each phase of Q-nMT formation requires a specific kinesin

(A) Images of representative WT cells (2 d) expressing Kar3-3GFP (green) and mTQZ-Tub1 (red) and corresponding fluorescence intensity along normalized Q-nMT bundles (see material and method section), bar is 2 µm.

(B) Morphometric Q-nMT bundle properties distribution in 4 d WT (grey), kar3Δ (red), vik1Δ (blue), and cik1Δ cells (green) expressing mTQZ-Tub1 after Noc treatment (30 µg/ml) – see Sup. Fig. 4A for the control without Noc. Each circle corresponds to an individual Q-nMT bundle. Black crosses are mean and SD. Images of representative cells are shown, bar is 2 µm.

(C) Nuclear MT length distribution in WT and cin8Δ cells expressing mTQZ-Tub1 treated (+) or not (-) with 15 min Noc (30 µg/ml). Each circle corresponds to the length of an individual MT structure, mean and SD are shown. A student test (t-test with two independent samples) was used to compare (+) or (-) Noc samples from the same experiment (n > 250, N = 2). The indicated p-values are the highest p-values calculated among the 2 repeated experiments. ***: p-value < 1.10−5.

(D) Fluorescence intensity along Q-nMT bundles in WT and cin8Δ cells expressing mTQZ-Tub1 7 h and 24 h after glucose exhaustion. Thin line: intensity from an individual cell; bold line: mean intensity (n > 60 / phase). Images of representative cells are shown, bar is 2 µm.

(E) WT and cin8Δ cells expressing mEOS3.2-Tub1 were imaged using PALM. Images in pseudo-colors of representative cells are shown. Full width at half maximum (FWHM) was measured at the indicated distance from the SPB. Each line in the bottom graphs corresponds to a single cell; p-value between WT and cin8Δ are indicated (unpaired t-test). Bar is 2 µm.

(F) Images of representative WT cells (3 d) expressing Kip1-GFP (green) and mTQZ-Tub1 (red). Graphs show fluorescence intensity along normalized Q-nMT bundles (plain and dash lines: mean and SD respectively). Bar is 2 µm, blue arrowhead: SPB.

(G) Representative images of WT, kip1Δ and kip3Δ cells expressing mTQZ-Tub1 treated (+) or not (-) with 15 min Noc (30 µg/ml), Bar is 2 µm.

(H) Nuclear MT length distribution in WT, kip1Δ and kip3Δ cells expressing mTQZ-Tub1 treated (+) or not (-) 15 min with Noc (30µg/ml). Legend is the same as in (C); n > 250; *: p-value < 1.10−3; ***: p-value < 1.10−6. MT mean length and SD are indicated.

Cin8 is a kinesin-5 that cross-links MTs (Bodrug et al., 2020; Pandey et al., 2021; Singh et al., 2018), and as such, could play a role in stabilizing Q-nMT bundles. Cin8 was barely detectable in quiescent cells (Sup. Fig. 4B) and cin8Δ cells assembled Q-nMT bundles that were thinner than in WT cells (Fig. 4C-E). However, these thinner bundles were stabilized in phase III (Fig. 4C and Sup. Fig. 4C-D). Taken together, these results strongly suggest that Cin8 is required for MT nucleation and/or elongation in phase II but not for stabilization in phase III.

Yeast have an additional kinesin-5 called Kip1 (Fridman et al., 2013). In quiescence, most of the Kip1-GFP signal was observed at the Q-nMT bundle +end (Fig. 4F). In kip1Δ cells, phase I was slightly delayed. In phase III, while as thick as WT Q-nMT bundles, MT bundles detected in kip1Δ cells were not fully stabilized (Fig. 4G-H and Sup. Fig. 4E). These data demonstrate that Kip1 is required for Q-nMT bundle stabilization in phase III. Of note, in phase I and II, Q-nMT bundles were slightly longer in cells deleted for the kinesin-8 Kip3, a MT depolymerase ((Fukuda et al., 2014), Fig. 4 G-H and Sup. Fig. 4G), but have a WT morphology in cells deleted for Kip2, a kinesin that stabilizes cytoplasmic MTs ((Hibbel et al., 2015); Sup. Fig. 4F-G). Taken together our results demonstrate that each phase of Q-nMT bundle assembly involves specific kinesins.

SPB duplication/separation requires Q-nMT bundle disassembly

Finally, we questioned the molecular mechanism of Q-nMT bundle disassembly upon exit from quiescence. We first found that cycloheximide prevented Q-nMT bundle disassembly indicating that this process requires de novo protein synthesis (Fig. 5A). Then we measured Q-nMT bundle thickness upon depolymerization, and found that while Q-nMT bundles shortened, they did not become thinner (Fig. 5B). In agreement, Nuf2-GFP clusters found at the Q-nMT bundle +end moved back to the SPB while the Nuf2-GFP clusters localized along the Q-nMT bundle remained immobile until they are reached by the +end-associated clusters (Sup. Fig. 5A). This shows that not all MT +ends started depolymerizing at the same time, and that longer MTs depolymerized first. This finding is consistent with our model in which MTs are cross-linked along the entire length of the Q-nMT bundle.

Q-nMT bundle disassembly always occurs before SPB separation upon quiescence exit

(A) WT cells expressing Spc42-RFP (red) and Nuf2-GFP (green) (5 d) were re-fed on an YPDA microscope pad. Individual Q-nMT bundles were measured in cells treated with DMSO (top panel, Nuf2-GFP, n=33) or with CHX (bottom panel, Stu2-GFP, n=49). Each line corresponds to an individual cell. In the upper panel, time was set to zero at the onset of MT bundle depolymerisation. In the lower panel, time was set to zero when cells were deposited on the agarose pad, bar is 2 µm.

(B) Q-nMT bundle length (green) and fluorescence intensity at full width half maximum (FWHM - orange) were measured upon quiescence exit in WT cells (5 d) expressing mTQZ-Tub1. Representative example of shrinking Q-nMT bundle is shown on the left, bar is 1 µm.

(C) Cells of the indicated genotype expressing mTQZ-Tub1 were grown for 4 d, and re-fed. Q-nMT bundle length was measured at the indicated time points, 15 min after a Noc treatment (30 µg/ml), to remove all dynamic cMTs. Each circle corresponds to a single cell. MT mean length and SD are indicated. A student test (t-test with two independent samples) was used to compare samples from the same experiment (n > 250, N = 4). The indicated p-values are the highest p-value calculated among the 4 repeated experiments. ***: p-value < 1.10−5. MT mean length and SD are indicated. Images of representative 4-day-old cells of the indicated genotype expressing mTQZ-Tub1 before and 60 min after quiescence exit are shown, bar is 2 µm.

(D) Representative images of a kip3Δ cell expressing Nuf2-GFP upon quiescence exit. Blue and white arrowheads: SPBs and Q-nMT bundle extremities respectively, bar is 2 µm.

(E) Percentage of 6-day-old cells expressing Spc42-mRFP1 with separated SPB as a function of time upon quiescence exit (n > 200).

(F) WT and kip3Δshe1Δ cells (6 d) expressing Spc42-mRFP1 (red) and mTQZ-Tub1 (green) were re-fed on a YPDA microscope pad. Percentage of cells with a single SBP with or without Q-nMT bundle or with duplicated SPBs were scored (N = 4, n > 200), SD are indicated. Images of representative cells are shown, bar is 2 µm. Right bottom panel: actin (phalloidin staining, red) in kip3Δ she1Δ cells (6 d) expressing mTQZ-Tub1 (green) before and 1 h after quiescence exit.

Since Kip3 is the only well-characterized yeast MT depolymerase, we examined Q-nMT bundle disassembly in kip3Δ cells and found that it was drastically delayed (Fig. 5C, Sup. Fig. 5B), yet the overall depolymerization rate remained unaffected (Sup. Fig. 5B, left panel). In fact, we found that in WT cells, Kip3 needs to be resynthesized upon exit from quiescence (Sup. Fig. 5C). We searched for proteins that might be involved in Q-nMT bundle disassembly together with Kip3. Among the proteins required for mitotic spindle disassembly, we found that inactivation of CDC14, CDC15, CDH1, DCC1, and TOR1 had no effect on Q-nMT bundle disassembly, as did the inactivation of Ipl1 or proteins involved in kinetochore-MT attachment (Sup. Fig. 5D-G). The only additional actor we found was She1, a dynein regulator (Bergman et al., 2012), whose deletion strongly exacerbated the kip3Δ phenotype (Fig. 5C, Sup. Fig. 5H). Thus, Q-nMT bundle disassembly does not rely on the canonical mitotic spindle disassembly machinery, but rather specifically requires Kip3 and She1.

Importantly, when we followed quiescence exit at the single cell level, depolymerization of the Q-nMT bundle always preceded SPB duplication/separation (Fig. 5A), in both WT and in kip3Δ cells (Sup. Fig. 5B, right panel). When we severely impeded Q-nMT bundle disassembly using kip3Δ she1Δ, we found that SPB separation was greatly delayed (Fig. 5D-F, Sup. Fig. 5I-K). Besides, the growth rate of kip3Δ she1Δ cells during proliferation is the same as WT cells (Sup. Fig. 5I), and upon quiescence exit, these mutant cells disassembled actin bodies as fast as WT cells (Sup. Fig. 5K), demonstrating that kip3Δ she1Δ cells were not impaired in sensing quiescence exit signals. These observations suggest that upon exit from quiescence, the Q-nMT bundle has to be disassembled prior to SPB duplication/separation.

Discussion

Our results shed light on the precise temporal sequences that generate a Q-nMT bundle upon quiescence establishment in yeast. From bacteria to human stem cells, changes in cell physicochemical properties are known to accompany the establishment of quiescence. Among these modifications, cellular volume reduction increases molecular crowding (Joyner et al., 2016), and pH acidification both increases viscosity and causes changes in macromolecule surface charges (Charruyer and Ghadially, 2018; Jacquel et al., 2021; Munder et al., 2016; Persson et al., 2020). Much evidence suggests that these modifications act as triggers for the auto-assembly of several types of enzyme-containing granules (Munder et al., 2016; Petrovska et al., 2014; Rabouille and Alberti, 2017), or complex structures such as P-bodies and Proteasome Storage Granules (Peters et al., 2013; Jacquel et al., 2021; Currie et al., 2023). Recently, Molines and colleagues demonstrated that cytoplasmic viscosity modulates MT dynamics in vivo (Molines et al., 2022). Here, we showed that the formation of the Q-nMT bundle does not depend on changes in the physicochemical properties that cells experience at the onset of quiescence establishment (Sup. Fig. 1). In fact, dynamic cMTs and a stable Q-nMT bundle can be observed simultaneously in a quiescent cell (Sup. Fig. 1I), while physicochemical properties evolve in parallel in both the nucleus and the cytoplasm (Joyner et al., 2016). Importantly, cellular volume reduction, pH acidification and increased viscosity are observed within minutes upon glucose starvation (Joyner et al., 2016) whereas the Q-nMT bundles require a couple of days to be fully assembled (Fig. 1). Furthermore, Q-nMT bundle formation cannot be delayed in time (Fig. 1H), and thus likely depends on a transient signal emitted upon glucose depletion. We speculate that the SPB may act as a platform that integrates this signal, and transfers it to the MT machinery in order to assemble a stable MT structure.

As it is required for the onset of Q-nMT bundle formation (Fig. 3B), the CPC component Aurora B/Ilp1, a kinase that plays a key role at the MT-kinetochore interface in response to the tension status, could be one of the targets of the aforementioned nutritional signal. In parallel, upon quiescence establishment, chromosome hyper-condensation (Guidi et al., 2015; Rutledge et al., 2015) could modify the tension at the kinetochore/MT interface and, as such, could contribute to the initiation of phase I (Fig. 6b). Besides, Bim1 has also been implicated in MT-kinetochore attachment (Dudziak et al., 2021; Akhmanova and Steinmetz, 2015). Although its deletion alone has no effect on Q-nMT bundle formation, its role becomes critical for phase I when kinetochore-MT interactions are already destabilized by the absence of other CPC components, other than Ipl1, such as Bir1 (Fig. 3D). Taken together, these observations suggest that kinetochore-MT interactions are essential for phase I initiation, as confirmed by the absence of Q-nMT bundle in the ndc80-1 mutant (Fig. 3A).

Model for Q-nMT bundle assembly

(a) In G1, the nucleus is in a Rabl-like configuration. (b) Upon quiescence establishment, chromosomes get condensed. MT-kinetochore interaction and Ilp1 are required for the onset of phase I. (c) Kar3 and its regulator Cik1 are essential to initiate Q-nMT bundle elongation. Although deletion of BIM1 has no effect, it becomes critical for phase I if kinetochore-MT interactions are destabilized by the absence of Chromosome Passenger Complex components. Kinetochore clustering by the monopolin complex and Slk19 is needed to maintain MT bundling while phase I MTs elongate. During phase I, Tub4 (ɣ-Tubulin) accumulates at the SPB. (d) In phase II, a second wave of MT nucleation and elongation occur. Phase II MTs are concurrently stabilized along pre-existing phase I MTs, in a Cin8-dependent manner. Phase I and phase II MT +ends (>1 µm) remain dynamic until the full-length Q-nMT bundle stabilization is reached via the action of Kip1, about 2 days after glucose exhaustion (e).

In phase I, SPB-anchored MTs elongate and are concomitantly stabilized (Fig. 1 A-C). This step requires the kinesin-14 Kar3 and its regulator Cik1 (Fig. 4B). During mating (Molodtsov et al., 2016), and in early mitosis, when half-spindles form (Kornakov et al., 2020), Kar3/Cik1, together with Bim1, align and cross-link growing MTs along existing MTs, thereby promoting the organization of MTs into parallel bundles. In mitosis however, the kinesin-14/EB1 complex promotes MT dynamics (Kornakov et al., 2020). It remains to be determined whether the robust MT stabilization in phase I depends on the modification of the kinesin-14/EB1 complex properties and/or on additional specific cross-linker(s) (Fig. 6c).

In quiescence, deletion of either Monopolin or SLK19 results in the formation of short, shattered and flared MT structures, phenotypes that are exacerbated by the deletion of BIM1 (Fig. 3F-H). Since Monopolin and Slk19 are involved in kinetochore clustering (Plowman et al., 2019; Rabitsch et al., 2003; Tóth et al., 2000; Lee et al., 2004; Mittal et al., 2019; Movshovich et al., 2008; Richmond et al., 2013; Zeng et al., 1999), we speculate that kinetochore-kinetochore interaction may not only help prevent depolymerization of individual MTs, but also constrain elongating phase I MTs and facilitate their concomitant cross-linking (Fig. 6d).

With the initiation of phase I, Tub4 begins to accumulate at the SPB to enable the second wave of MT nucleation observed in phase II. As phase II MTs elongate, they are simultaneously stabilized along pre-existing phase I MTs (Fig. 6d). This step relies on the kinesin-5/Cin8 as Q-nMT bundles assembled in cin8Δ cells are thinner than WT Q-nMT bundles (Fig. 4C-E). Intriguingly, in vitro, the monomeric form of the vertebrate kinesin-5, Eg5, promotes MT nucleation and stabilizes lateral tubulin-tubulin contacts (Chen et al 2019). We assume that in the absence of Cin8, either nucleation of phase II MTs cannot occur, or phase II MTs can elongate, but cannot be cross-linked and stabilized to a pre-existing phase I MT(s) as they grow, and thus rapidly depolymerize. Absence of bundle thickening in phase II is also observed in the “Tub1-only” mutant i.e. in the absence of Tub3. Tubulin isotypes or tubulin PTMs are known to promote the recruitment of specific kinesins and thereby modify MT properties, see for example (Sirajuddin et al., 2014; Peris et al., 2009; Dunn et al., 2008). Could Tub3 be important for Cin8 recruitment or function during phase II?

Following phase II, MTs elongate, but their distal parts (>1 µm) are not yet stable (Fig. 1A). Stabilization of full-length Q-nMT bundles is achieved approximately 4 days after glucose exhaustion. This step depends on the kinesin-5/Kip1 (Fig. 4G-H and Sup. Fig. 4E) which is essential for cross-linking and stabilizing elongating MTs during phase III (Fig. 6e). Accordingly, we observed a Kip1 enrichment at the Q-nMT bundle +ends (Fig. 4F). Thus, consistent with their ability to form homo-tetramers capable of cross-linking (Acar et al., 2013; Kapitein et al., 2005; Scholey et al., 2014; Weinger et al., 2011) and stabilizing parallel MTs (Kapitein et al., 2005; Shimamoto et al., 2015; Yukawa et al., 2020) in addition to their role of sliding anti-parallel MTs, the two yeast kinesins 5 are essential for Q-nMT bundle formation. However, in quiescence, as well as in mitosis, Kip1 and Cin8 have non-equivalent functions (Roostalu et al., 2011; Shapira and Gheber, 2016).

Thus far, our results indicate that the mechanism of Q-nMT bundle formation is distinct from that of mitotic spindle assembly. Similarly, we show that Q-nMT bundle disassembly does not involve the same pathway as mitotic spindle disassembly as it involves Kip3 (Fig. 5C) but is independent of the anaphase-promoting complex and Aurora B/Ipl1 (Sup. Fig. 5D-G). Interestingly, the longest MTs begin to depolymerize from their +ends first. Then, when they reach the +ends of shorter MTs, the shorter ones proceed to depolymerize alongside them (Sup. Fig. 5A). This cooperative behavior has been observed in vitro within parallel MT bundles (Laan et al., 2008). Our data demonstrate that such a comportment does exist in vivo.

Finally, our data indicate that the Q-nMT bundle is required for the survival of WT cells in quiescence. While we do not yet know why this structure is important for chronological aging, it is clear that the presence of the Q-nMT bundle modifies the organization of the nucleus (Laporte et al., 2013; Laporte and Sagot, 2014; Laporte et al., 2016). Since chromatin organization influences gene expression, one hypothesis could be that the presence of the Q-nMT bundle is required for the expression of genes necessary for cell survival in quiescence. Another attractive hypothesis is that the Q-nMT bundle could be the yeast analog of the mammalian primary cilium. Indeed, these two structures are formed upon proliferation cessation, they are both composed of highly stable parallel MTs, their formation involves MT motors, and they are both templated from the centriole/SPB. More importantly, we show here that Q-nMT disassembly is required to allow SPB separation upon exit from quiescence (Fig. 5). Thus, the Q-nMT bundle disassembly may control reentry into the proliferation cycle, just as the primary cilium does in ciliated mammalian cells (Goto et al., 2017; Kim and Tsiokas, 2011).

Materiel and Methods

Yeast strains, plasmids and growth conditions

All the strains used in this study are isogenic to BY4741 (mat a, leu2Δ0, his3Δ0, ura3Δ0, met15Δ0) or BY4742 (mat alpha, leu2Δ0, his3Δ0, ura3Δ0, lys2Δ0), available from GE Healthcare Dharmacon Inc. (UK), except for Fig. 2, in which strains of the S288c background were used (Nsamba et al., 2021) and Sup. Fig. 1F in which the W303 background was used. BY strains carrying GFP fusions were obtained from ThermoFisher Scientific (Waltham, MA, USA). Integrative plasmids pTub1-mTurquoise2-Tub1, pTub1-wasabi-Tub1, pTub1-mRUBY2-Tub1, pTub1-mEOS2-Tub1 were a generous gift of Wei-Lih Lee (Markus et al., 2015). The strain expressing SPC42-mRFP1 is a generous gift from E. O’Shea (Huh et al., 2003). The strain expressing TUB4-mTQZ is a generous gift of S. Jaspersen (Burns et al., 2015). The Ipl1-5as is a generous gift from A. Marston. Ipl1-1, ndc80-1 and Sli15-3 are a generous gift from C. Boone. Three copies of eGFP in tandem were integrated at the 3’ end of DAD2, TUB3 or KAR3 endogenous loci respectively. Plasmids for expressing Nup2-RFP (p3695) or Bim1-3xGFP (p4587) from the endogenous locus were described in (Laporte et al., 2013).

For Fig. 1F, pHIS3:mTurquoise2-Tub1+3’UTR::LEU2 was integrated at the TUB1 locus, then a pRS303-ADH2p-mRuby2-Tub1 was integrated at the HIS3 locus. To generate this plasmid, the ADH2 promoter was amplified from yeast genomic DNA flanked with NotI and SpeI restrictions sites and inserted in pRS303. The mRuby2-Tub1+3’UTR, including the 116-nucleotide intron and 618 nucleotides downstream of the stop codon was cloned between SpeI/SalI sites.

Yeast cells were grown in liquid YPDA medium at 30 °C in flasks as described previously in (Sagot et al., 2006) except for experiments using thermosensitive strains, where cells were first grown at 25 °C then shifted at 37 °C for the indicated time before imaging.

The experiment in Sup. Fig. 1J was performed as described in (Orij et al., 2009; Peters et al., 2013). In brief, proliferating cells were transferred at an OD600nm of 0.5 in Hepes buffer (25 mM Hepes, pH 7.4, 200 mM KCl, 1 mM CaCl2, and 2 % dextrose), buffered either at pH 4 or 7.5, in the presence of 100 µM CCCP (Sigma-Aldrich). After 150 min 30 °C shaking, cells were imaged (for mTQZ-Tub1) or fixed and stained using Alexa Fluor 568–phalloidin (Invitrogen) as described (Sagot et al., 2006).

For live cell imaging, 2 μL of the cell culture were spotted onto a glass slide and immediately imaged at room temperature.

For quiescence exit, cells were first incubated 2-3 min in liquid YPD and then 2 µL were spread onto a 2 % agarose microscope pad containing YPD. Individual cells were imaged every hour, up to 6 or 12 h at 21°C. For quiescence exit in presence of cycloheximide (Fig. 5A), cells were pre-incubated 30 min in the presence of the drug prior to quiescence exit. Cycloheximide was used at 180 µM (Sigma-Aldrich), Nocodazole was used at 30 µg/mL (7.5 µM) (Sigma-Aldrich), NA-PP1 was used at 50µM (Sigma-Aldrich).

Fluorescence Microscopy

Cells were observed on a fully-automated Zeiss 200M inverted microscope (Carl Zeiss, Thornwood, NY, USA) equipped with an MS-2000 stage (Applied Scientific Instrumentation, Eugene, OR, USA), a Lambda LS 300 Watt xenon light source (Sutter, Novato, CA, USA), a 100X 1.4NA Plan-Apochromat objective, and a 5 position filter turret. For RFP and mRuby2 imaging we used a Cy3 filter (Ex: HQ535/50 nm – Em: HQ610/75 nm – BS: Q565 nm lp). For GFP imaging, we used a FITC filter (excitation, HQ487/25 nm; emission, HQ535/40 nm; beam splitter, Q505 nm lp). For mTurquoise2 imaging we used a DAPI filter (Ex: 360/40 nm – Em: 460/50 nm– BS: 400 nm). All the filters are from Chroma Technology Corp. Images were acquired using a CoolSnap HQ camera (Roper Scientific, Tucson, AZ, USA). The microscope, camera, and shutters (Uniblitz, Rochester, NY, USA) were controlled by SlideBook software 5.0. (Intelligent Imaging Innovations, Denver, CO, USA).

For PALM (Fig. 4E) a Nikon Ti-Eclipse equipped with iLas2 TIRF arm, laser diodes (405, 488, 532, 561, 642 nm), a 100x 1.49 oil (TIRF) objective connected to a EMCCD Camera Photometrics Evolve was used.

Immunofluorescence was done as in (Laporte et al., 2016) and actin AlexaFluor Phalloidin (Invitrogen) staining as in (Sagot et al., 2006).

For expansion microscopy (Sup Fig. 2A and 3A), spheroplasts were obtained as described in (Laporte et al., 2016). After washing in 10 mg/mL NaBH4 in PEMS for 10 min, spheroplasts were seeded on 12 mm round poly-L-lysine coated coverslips, washed twice for 30 min in 100 μL PEM-BAL (PEM + 1% BSA) and incubated with the primary antibodies (anti-GFP from mouse; Roche, ref. 11814460001, 1/50; anti-alpha-tubulin from rat; YOL1/34, Abcam, ref. Ab6161, 1/100) for 1 hour at 37°C. Cells were then washed three times in PEM-BAL and incubated with the secondary antibodies (Goat anti-mouse AlexaFluor®488 1/200 and Donkey anti-rat AlexaFluor®555; A11029 and A21434 respectively, ThermoFisher Scientific, Waltham, MA) for 45 min at 37 °C. Cells were then washed three times with PBS. Cells were processed for Expansion Microscopy as previously described in (Bahri et al., 2021). Spheroplasts were incubated for 10 min in 0.25 % GA in PBS, washed in PBS three times for 5 min, and processed for gelation. A drop of 100 µL of ExM MS (8.625% (wt/wt) SA, 2.5% (wt/wt) AA, 0.15 % (wt/wt) BIS, 2 M NaCl in 1× PBS) was placed on the chilled parafilm, and coverslips were put on the drop with cells facing the solution and incubated for 3 min. Then the coverslips were transferred to 35 μL of ExM MS supplemented with 0.2 % APS and 0.2 % TEMED, with the initiator (APS) added last. Gelation proceeded for 3 min on ice, and samples were incubated at 37 °C in a humidified chamber in the dark for 1 h. Then coverslips with attached gels were transferred into a six-well plate for incubation in 2 mL of digestion buffer (1× TAE buffer, 0.5% Triton X-100, 0.8 M guanidine hydrochloride, pH ∼8.3) supplemented with DAPI (1 µg/mL) for 10-15 min at 37 °C, until the gels detached. Fresh proteinase K at 8 units/mL was then added and samples incubated at 37 °C for 30 min. Finally, gels were removed and placed in 10mL petri dishes filled with ddH2O for expansion. Water was exchanged at least twice every 30 min, and incubated in ddH2O overnight at RT. Gels expanded between 4× and 4.2× according to SA purity. Expanded cells were imaged with a UPlanS Apo 100×/1.4 oil immersion objective in an Olympus IX81 microscope (Olympus, Tokyo, Japan). For structured illumination microscopy (Sup Fig. 3A), a ZEISS Elyra 7 Lattice SIM was used.

Laser ablation (Fig. 3I and Sup. Fig. 3K-L) was performed at room temperature with a 100X oil Plan-Apochromat objective lens (NA 1.4) and an Axio-Observed.Z1 microscope (Carl Zeiss) equipped with a spinning disk confocal (Yokogawa), an EMCCD Evolve camera (Photometrics and Roper Scientific) and 491 nm (100 mW; Cobolt calypso) and 561 nm (100 mW; Cobolt Jive) lasers. Images were acquired with Metamorph software (Roper Scientific). Every 30 to 120 s, a Z-series of 0.4 µm steps were acquired. A 355 nm microchip laser (Teem Photonics) with a 21 kHz repetition rate, 0.8 µJ energy/pulse, 2 kW of peak power and 400 ps pulse width, powered with an iLas2 PULSE system (Roper Scientific) was used between 10 to 40 % power with one pulse of a spot length of 100 points. Breakage was considered as successful if a non-alignment between the two remaining Q-n MT bundle fragments was observed.

In Fig.1H, F; Fig. 2B-C; Fig.3C; Fig. 4D, G; Sup. Fig. 1I; Sup. Fig. 2E and Sup. Fig. 3A, a faint and fuzzy fluorescence signal was detected in some cell cytoplasm using the GFP filter set. This signal is not GFP but rather due to a non-specific yellow background fluorescence.

Z-stacks were deconvolved using the Deconvolution Lab plugin (Fig. 1D, F; Fig. 3 C, E-F Fig. 4A, F; Sup. Fig. 1E; Sup. Fig. 2A and Sup. Fig. 4B).

Image analysis

Distribution and associated statistics were performed using GraphPad Prism 5 (GraphPad Software, Inc. La Jolla, USA) and Excel (Microsoft). Unless specified, a student test (t-test with two independent samples) was used to compare two conditions. Among p-values obtained from biological repetitions in a same experiment (Fig.1A; Fig. 2 B-C; Fig. 3D, F, H; Fig. 4 C, H, and Fig. 5C), the indicated p-values are the highest measured among repeated experiments. P-values above 0.05 are indicated, while values below are depicted with ** (0.05>p-value>1.10−4) and *** if p-value is less than 1.10−5 Histograms and scatter dot plots show the mean and the error bars indicate SD. MT length was measured on MAX-projected images using image-J.

For MT fluorescence intensity measurements (Fig. 1B, Fig. 2D and Fig. 4D) a line scan (i1) of 4 pixels width containing both FP signal and background was drawn along MTs on SUM projection images (4 z-plans) using ImageJ. A line of 8 pixels width (at the same location) was drawn in order to calculate the intensity of the surrounding background (i2). The real intensity (ir) was calculated as follow: int = [i1 - ([(i2 x i2surface) - (i1 x i1surface)] / (i2surface - i1surface)) * i1surface]. We arbitrarily set the ‘zero’ at 0,5µm before the fluorescence increased onset on the SPB side in order to align all intensity measurements.

A similar approach was conducted to measure Tub4 fluorescence intensity (Fig. 1E, Sup. Fig. 1F) i.e. (i1) and (i2) were two boxes, the second including the first one. Intensity was calculated as follow: int = [i1 - ([(i2 x i2surface) - (i1 x i1surface)] / (i2surface - i1surface)) * i1surface].

For morphometric Q-nMT bundle property distribution (Fig. 1C and 4B) the mean fluorescence intensity was measured in each individual cell, in a finite area localized adjacent to the SPB (0,4 µm – 3-4 Z stack sum projected) as an estimate of Q-nMT bundle width and plotted as a function of the according measured bundle length.

For Nuf2-GFP fluorescence intensity measurement at the SPB (Sup. Fig. 1E, bottom panel), GFP and RFP line scan measurements were done on a 4 Z-planes sum projection. The “SPB localization zone” was defined as the length ending 250 µm after the brightest RFP signal of Spc42-RFP. The remaining Q-nMT bundle zone was defined as the “+end zone”. The total Nuf2-GFP signal detected along the Q-nMT bundle was set to 100 % to determine the percentage of the signal measured at the “SPB localization zone” and the “+end zone”.

Normalized Q-nMT bundle length (Fig. 3C) was done after line scan intensity measurements. mTQZ-Tub1 intensity slopes were first aligned on their inflexion points. To compare Q-nMT bundles with different lengths, we first sorted Q-nMT bundles <1.8 µm. After an artificial isotropic expansion, we fit all MT structures to length order. Then, the corresponding mean intensity of the Q-nMT bundle (mTZQ-Tub1) and GFP signal were calculated.

To measure Q-nMT bundle depolymerization (Fig. 5A and Sup. Fig. 5B), individual Q-nMT bundle lengths were measured over time. The first length measurement was set to 100%. A fluorescence drop above or equal to 25 % defines the inflexion point of the slope, and was used to align the length measurements of different Q-nMT.

The FWHM (Full Width at Half Maximum, Fig. 5B) was calculated by measuring fluorescence intensity of a line crossing perpendicularly to a Q-nMT bundle at 0.5 µm from the SPB, using a sum projection. After fitting the intensity level with a Gaussian distribution and obtaining the standard deviation (σ) value, FWHM was calculated using the equation 2√(2ln2)xσ.

Western blots

Western blots were done as described in (Sagot et al., 2006) using anti-GFP antibodies (Abcam); anti-Tat1 antibodies, a generous gift from J-P. Javerzat; antibody against the budding yeast Act1 or Sac6, generous gifts from B. Goode; and anti-TagRFP (mRuby2) antibodies, a generous gift from M. Rojo.

Phenotypical analysis of cells without Q-nMT bundle

WT strains were grown to glucose exhaustion (OD ∼6.5, Fig. 1F) or to 5 d (Sup. Fig. 1L), washed-out with “old YPDA” (YPDA medium in which cells were grown for 4 d and then filtered to remove cells) pulsed with either DMSO alone or DMSO containing Noc (30 µg/mL final concentration) for 24 h. Cells were then washed twice in “old YPDA” before being incubated in the same medium. Dead cell measurement was done using methylene blue staining. The capacity of cells to exit quiescence (Sup. Fig. 1L, right panel) was scored after micro-manipulation of live cells (cells not stained with methylene blue) as described in (Laporte et al., 2011).

Supplemental data

Supplementary Figure 1 further describes the 3 steps of Q-nMT bundle formation. It shows the impact of physicochemical cell properties on MT stabilization and demonstrates that the Q-nMT bundle is required for both cell viability in quiescence and quiescence exit fitness.

Supplementary Figure 2 describes the impact of the alpha tubulin level on Q-nMT bundle assembly.

Supplementary Figure 3 demonstrates that kinetochore-MT interactions are not required for Q-nMT bundle maintenance and that mutants affected for kinetochore-kinetochore interactions have reduced viability in quiescence.

Supplementary Figure 4 further describes the impact of kinesin deletion on Q-nMT bundle morphometric parameters.

Supplementary Figure 5 explores Q-nMT bundle disassembly in mutants involved in mitotic spindle dismantlement.

Supporting information